Synthetic Eco-Evolutionary Dynamics in Simple Molecular Environment

Version 1

Abstract

Abstract

The understanding of eco-evolutionary dynamics, and in particular the mechanism of emergence of species, is still fragmentary and in need of test bench model systems. To this aim we developed a variant of SELEX in-vitro selection to study the evolution of a population of ∼ 1015 single-strand DNA oligonucleotide ‘individuals’. We begin with a seed of random sequences which we select via affinity capture from ∼ 1012 DNA oligomers of fixed sequence (‘resources’) over which they compete. At each cycle (‘generation’), the ecosystem is replenished via PCR amplification of survivors. Massive parallel sequencing indicates that across generations the variety of sequences (‘species’) drastically decreases, while some of them become populous and dominate the ecosystem. The simplicity of our approach, in which survival is granted by hybridization, enables a quantitative investigation of fitness through a statistical analysis of binding energies. We find that the strength of individual-resource binding dominates the selection in the first generations, while inter and intra-individual interactions becomes important in later stages, in parallel with the emergence of prototypical forms of mutualism and parasitism.

Introduction

A central effort in theoretical biology and ecology is to provide an effective description of the intimate but often subtle relationship between a given environment and the evolution of its ecosystem. In the case of simple environments without geographical isolation and physical barriers, as the one here considered, the emergence of species and phenotyping clustering Davis (1943); Rundle and Nosil (2005); Keymer et al. (2012); Gupta et al. (2021) is generally considered an outcome of the competition for the limited available resources Dieckmann and Doebeli (1999); Pigolotti et al. (2007); de Aguiar et al. (2009); Anceschi et al. (2018). The coexistence of stable species is ecologically understood in terms of “niches”, indicating the unique role and position that a particular species occupies within an ecosystem. According to the “niche hypothesis” Chase and Leibold (2009); Peterson (2011); Anceschi et al. (2018), biodiversity is limited by the number of “niches” (or types of resources) that are present since no two species can occupy the same niche indefinitely, as one would eventually out-compete the other Levin (1970); Gupta et al. (2021).

Fitness, which quantifies species reproductive success and thus also their relationship with the environment De Visser and Krug (2014), cannot generally be defined in a predictive way even in the most idealized systems Thurner et al. (2010); Wiser and Lenski (2015). Indeed, despite the efforts to identify simple case study systems, the evolution of populations formed by a variety of species remains difficult to model and to quantitatively characterize because of the inherent complexity of ecosystems and living beings, of the large number of potentially relevant variables of difficult access, and of the role of stochasticity Catalán et al. (2017).

In this scenario, introducing new tools to explore and test eco-evolutionary models, concepts and interpretations appear as the best strategy toward new understanding Solé (2016). Along this line, various synthetic biological platforms have been proposed in the last years Ichihashi et al. (2013); Tizei et al. (2016); Parrilla-Gutierrez et al. (2017); Kauffman et al. (2018); Adamala and Szostak (2013); Katla et al. (2023) that exploit different principles and mechanisms, and focused on in-vivo, in-vitro, ex-vivo, or in-silico approaches. For example, in Ichihashi et al. (2013) a “cell-like” model system is proposed, in which the evolution of one long genomic RNA (> 2,000 nt-long) was investigated in detail under the action of a selective pressure ultimately provided by its biological meaning. Despite the tremendous simplification provided by this approach, the constructed artificial cell is still a complex system that includes ribosomes, lipids, translation factors, accessory proteins, tRNAs and amino acids. While most of the synthetic biological platforms are based on the establishment of cell-like, compartmentalized systems involving complex biomolecular milieus and processes Tizei et al. (2016), a few are devoted to investigate, with distinct strategies, the competition between two coexisting vescicle-based species competing for resources (vescicle-forming molecules) in a non-biochemical context Adamala and Szostak (2013); Katla et al. (2023). These articles demonstrate that typical evolution concepts, such as competition or resources and niche exclusion principle, can be applied to “non-biological” systems.

We here propose a synthetic eco-evolutionary scheme, with no cell-like compartmentalization and no connection to the molecular biology of the cell: no coding sequences, no translation, no proteins involved. We consider an ecosystem formed by a large crowd (∼ 1015) of distinct molecular individuals interacting with each other and competing for survival in an environment with fixed resources, and we focus on the process by which selection and competition drive the emergence of dominating species. Specifically, we study solutions of 50-base-long single strand DNA oligomers with random sequences, thus initially all distinct, whose mechanisms of survival and mutual interactions are based on DNA hybridization. By exploiting the wealth of tools and knowledge around DNA code selective pairing and DNA synthesis, amplification and sequencing, we create a condition in which a limited number of accessible and computable variables controls the destiny of an ecosystem formed by biological molecules that evolve in a non-biological way, i.e. with no reference to the biological meaning of their sequence.

The results of our work, demonstrating the intimate connection between fitness and ecological interactions, belong to the growing body of studies Fussmann et al. (2007); Vetsigian (2017); Camacho Mateu et al. (2021) showing that ecological and evolutionary processes are strictly connected in the emergence and maintenance of species.

Results

Affinity-based DNA Synthetic Evolution

We introduce here a variant of SELEX for in-vitro synthetic evolution of oligonucleotides to develop protein-binding aptamers Ellington and Szostak (1990); Tuerk and Gold (1990). In standard SELEX protocol, the evolving oligonucleotides are selected at each cycle by their interactions of with the target protein. In the experiment here reported, we implemented a selective mechanism based on the affinity capture provided by magnetic beads carrying single-stranded DNA (ssDNA) filaments of fixed length L = 20 and sequence, that act as targets (or resources, Figure1a). Selection is thus primarily based on the sequence of the DNA individuals and its level of complementarity to the targets. This marks a significant difference with SELEX, in which the aptamer-protein interaction depends instead on higher order factors such as the secondary structure of the oligonucleotides and the variety of binding sites on the folded protein. Being these hard to model, predict and control, SELEX has never been considered, to the best of our knowledge, as a useful experimental test bench to understand evolution.

In our Affinity-based DNA Synthetic Evolution (ADSE) protocol, evolution start from an initial pool of DNA Individuals (DNAi), chosen to be of fixed length L = 50 and random-sequence. Since the potential molecular variety is 450 ∼ 1030 while our experiments use about 1015 initial molecules, each sequence is represented one time only. The following evolution process is sketched in Figure1b, Figure 1-figure supplement 2, and is given by three steps.

Affinity-based DNA Synthetic Evolution (ADSE). a: structure of the DNA oligomers participating in ADSE as individuals (DNAi) and resources (Figure 1-figure supplement 1). b: steps in the ADSE (Figure 1-figure supplement 2). The process starts with a random-sequence DNAi population. The capture by magnetic bead-conjugated resources provides the selection: bead-bound DNAi are amplified to form the new generation, a small fraction of which is sequenced by MPS. The rest of the original solution is discarded. Red arrows mark the steps of each ADSE cycle. c: possible interactions involving DNAi. The online version of this article includes the following figure supplement(s) for Figure 1: Figure 1-figure supplement 1 DNA individuals structure and its interaction with the oligos. Figure 1-figure supplement 2 Schematic representation of the Affinity-based DNA Synthetic Evolution protocol.

Selection: the seed population is mixed with a given amount of dispersed capture beads. After a suitable incubation time, the beads are extracted from the solution and the bound oligomers released and saved. The rest of the original solution is discarded;

Amplification: the pool of “survived” oligomers is PCR amplified about 1000 times to recover the initial molarity;

Sequencing: a small portion of the amplified sample (1-3 106 molecules) is analyzed by massive parallel sequencing.

These steps constitute a cycle - one generation of evolution - that we repeated up to 24 times in two independent evolution histories.

In actuality, to enable amplification and sequencing, DNAi are built by flanking the 50mer with two 25 base long fixed sequences that enable primer binding for a total length of 100 bases. Such two terminal segments of the DNAi are made inactive during selection by hybridization with oligomers of perfect complementarity, as sketched in Figure 1a, Figure 1-figure supplement 1.

A key feature of ADSE is that DNAi can interact not only with the resources, but also with itself and with each other, and also potentially form complexes binding to the affinity beads as summarized in Figure 1c. Indeed, the choice of a length of 50 for the DNAi interaction was thought to enable - in principle - simultaneous resource and mutual binding.

PCR amplification can be of high fidelity or error-prone, the latter choice enabling genetic drift and, potentially, speciation. Since our primary goal was to first investigate fitness in a pool of competing species within a given niche - defined in our case by the 20 base long resources, we opted for a high fidelity amplification, and left the investigation of high mutations regimes to a follow-up experiment. Because of the many PCR cycles required in the ADSE scheme, the amplification inevitably leads to the formation of artifacts in the form of longer sequences Tolle et al. (2014), a phenomenon that intrinsically sets a limit to the number of generations that can be explored (see Appendix 1).

As detailed below, the ADSE protocol enables observing a non-trivial evolution of the DNAi ecosystem across generations, the emergence of dominating species and nonmonotonic population evolution. It also enables to appreciate the role of inter-specific interactions and their contribution to fitness.

Evolution of the DNAi ecosystem

The main output of the synthetic evolution that we are proposing is the dataset {DNAi}j obtained by sequencing DNAi at the various cycles (1 ≤ j ≤ 24 being the index of the cycle). We find that the initial random ecosystem markedly changes with ADSE generations as shown in Figure 2a, in which we have plotted the evolution of: (i) the fraction FD of the total population formed by distinct nucleotide sequences (red dots). FD drops from 1 to nearly 0, indicating that initially DNAi are all different from each other, while after 24 generations the number of distinct sequences is much smaller than the number of DNAi. This indicates that most of the initial sequences become progressively “extinct”; (ii) The fraction F10 of the total population formed by the 10 most abundant DNA sequences (black dots). F10 is initially close to 0 - being each sequence represented by a single DNAi, and grows to about 25% at cycle 12 and 55% at cycle 24 - indicating that the offspring of 10 initial “mother” sequences becomes the majority of the system. This loss of diversity can also be quantified by the zipped file size of the list of sequences Benedetto et al. (2002) or by computing its Shannon entropy Hill (1973), both markedly decreasing during evolution (see Appendix 1).

Evolution of the DNAi population. Time is expressed in ADSE cycles. a: fraction FD of the total population formed by different sequences obtained from experiment (red dots) and computed with the IBEE model (green line); fraction F10 of the total population formed by the 10 most abundant sequences (experimental: black dots, IBEE model: grey line). c: boxplots and scatterplots of ⟨ ΔGDR⟩ in ensembles of 1000 random sequences (left), 1000 sequences extracted from the experimental population at cycle 24 (middle) and top-10 most populous sequences at the same cycle (right). The color code is assigned to each point based on its ω value (color bar). d-f: probability distributions P (ω) for at cycles 0 (d), 12 (e) and 24 (f). The orange points and line are the distributions evaluated with the null model. g-i: evolution of the abundance (expressed as fraction of the total population F) of sequences whose ω is low (3 ≤ ω ≤ 5-panel g), medium (6 ≤ ω ≤ 10 - panel h) and high (11 ≤ ω - panel i) as obtained from the experiments (dots) and with the IBEE model (green lines). Model results are averages over 20 simulations. The online version of this article includes the following figure supplement(s) for Figure 2: Figure 2-figure supplement 1. Time evolution of the fraction of PCR by-products and of their ⟨ω⟩. Figure 2-figure supplement 2. Probability distributions P (⟨ω⟩) of PCR by-products at cycles 0 and 24. Figure 2-figure supplement 3. Time evolution of ⟨ ω ⟩for the replica Oligo2. Figure 2-figure supplement 4. Probability distributions P (⟨ω⟩) at cycles 0 and 18 for the replica Oligo2. Figure 2-figure supplement 5. Experimental vs IBEE time evolution of the population zip ratio and of the Shannon Entropy associated to the RSA distribution. Figure 2-figure supplement 6. Experimental vs IBEE ⟨ω⟩ time evolution, for two different IBEE hyperparameters choices. Figure 2-figure supplement 7. Experimental vs IBEE ⟨ω⟩ time evolution, for different IBEE starting population sizes and compositions.

Since in ADSE survival depends on bead capture, we expect the DNAi-resource binding strength to be an essential ingredient of fitness. To this aim we computed the mean DNAi-resources binding free energy ΔGDR across generations, plotted in Figure2b (blue squares, right-hand side y-axis). Specifically, ΔGDR has been computed based on DNAi and resource sequences by using the standard “nearest-neighbor approximation” for the thermodynamics of DNA hybridization SantaLucia Jr (1998); Ghosh et al. (2019); Plata et al. (2021), as implemented in the NUPACK tool in Python Zadeh et al. (2011). Being this computation of some complexity, we could perform it only on batches of 1000 DNAi randomly chosen within {DNAi}j, the error bars expressing the uncertainty introduced by such downsampling (see Material and Methods section).

To perform a faster - and thus complete - analysis of the resource-binding energy in the evolving population, we introduced a simpler, albeit more approximate, quantifier: the length ω of the longest consecutive number of basis within each DNAi that is complementary to the resource sequence (considering all possible relative positions of the two), in which we allow for single pairing errors and 1-base bulges Mambretti et al. (2022a). While ⟨ω⟩, missing key ingredients such as the fraction of CG vs. AT base pairs and their sequence, is certainly inadequate to evaluate the binding strength of a specific DNAi, we find that its average value ⟨ω⟩computed for the whole population in each generation (Figure2b, full red dots, left-hand side y-axis) grows almost identically to ⟨ΔGDR⟩. The ⟨ω⟩ axis has been scaled so to make its value computed on a pool of ideal randomly sequence DNAi (dashed red line) match, on the plot, ⟨ΔGDR⟩ computed on the same population (dashed blue line). A more detailed comparison between ω and ΔGDR is in Figure2c, showing box plots with ΔGDR on the y axis and ω expressed through color code, built on a random selection of 1000 distinct DNAi in the initial (left hand side box) and the final populations (central box), and on top 10 most frequent sequences in the final generation (right hand side box). The result further strengthens the validity, in our statistical context, of ω as a quantifier of interaction strength to the resources Mambretti et al. (2022a).

Figure2b reveals that, during ADSE, ω tends to saturate at a value ωsat ≈ 10, indicating that no relevant selective advantage is gained when ω > 10, in agreement with the notion that the residence time of hybridized oligomers becomes larger than typical experimental times when ω > 12 Di Leo et al. (2022).

Figure 2d-f describes the evolution of the ecosystem {DNAij by showing P (ω), the fraction of DNAi having overlap ω with the resource evaluated in the initial pool (panel d) and at generation 12 (e) and 24 (f). P (ω) clearly evolves, its small ω components being progressively lost, while individuals with large ω grow in number, as they are more successful in being selected and amplified. It might be worth pointing out that the appearance of non-zero p(ω > 13) at generation 12, while p(ω > 13) = 0 in the initial distribution, is a pure effect of the undersampling involved in the sequencing procedure.

The different destiny of DNAi with distinct ω values is shown in Figure 2g-i, where we show the temporal (i.e. across generation) evolution of the fraction of the total population whose ω is in the following ranges: 3 ≤ ω ≤ 5 (panel g), 6 ≤ ω ≤ 10 (panel h) and 11 ≤ ω (panel i). As expected, DNAi with weak affinity to the resources decrease while those with large affinity increase. Interestingly, DNAi with intermediate affinity exhibit a non-monotonic behavior, indicating that the conditions for survival change during the evolution, reflecting the evolution of the ecosystem.

Null Model and Eco-evolutionary Algorithm

In order to better understand the experimental outcomes, we first build a null random model without evolution and then an Individual-Based Eco-Evolutionary (IBEE) model enabling predictions for P (ω).

The null model describes the interaction between individuals in the initial pool {DNAij=0} of random sequences and the resources within a purely combinatorial framework (see Methods and Appendix 3 for mathematical details). In this model, we attach a random string of length 50 to the resource string in a random position, and we compute the maximum consecutive overlap, accepting the binding only if it is at least formed by 3 complementary basis. For analytical tractability we do not allow the random string to shift over the resource, but we calculate the ω constrained to a specific binding position.

The resulting analytical p0(ω), plotted in Figure 2d (orange crosses) closely matches the data obtained from sequencing, confirming the random-sequence nature of {DNAij=0}. The analytical P0(ω) extends to a ω range where no data are available because of the limited size of the sequenced pool. The comparison between P0(ω) and P (ω) in panels e and f shows that the exponential decay of P0(ω) at large ω is maintained even in generation 12 and partly in generation 24, further supporting the notion that the selective advantage of ω saturates at large ω.

In the IBEE algorithm we consider Np individuals. Each has a fitness fi(ω) (i = 1, 2, .., Np) that depends on its affinity ω with the resource. At time t = 0 we assign ω to each individual based on P0(ω) from the null model. Then, for each evolutive cycle we model competition and selection so that out of Np individuals, only Nr survives. This is attained as a combination of two processes: a fraction x of the Nr sequences results from sampling individuals from the Np population, each with a survival probability f (ω) ∈ (0, 1); the remaining fraction 1 − x is formed by a straight random selection from the Np individuals. The latter group is meant to mimic “neutral drift” (as it would be expressed in evolutionary language) provided by non-specific binding to the beads. The Nr survivers are amplified by identical copying back to Np, the starting population of the next evolutionary cycle (the introduction of very rare mutations do not affect the results). We perform 24 evolutionary cycles.

As expected, the outcome of the eco-evolutionary dynamics strongly depends on the shape of the fitness function. We model it as: where ωmax is the (cycle-dependent) largest value of ω within the actual population and, ω* = ω if ω < ωsat, while ω* = ωsat otherwise. ωsat and γ are parameters to be tuned in the comparison with the observed⟨ω⟩. ωsat expresses the loss of fitness gain for ω > ωsat yielding the saturation of⟨ω⟩. γ represents the strength with which f (ω) depends on ω, and thus the rate at which low ω individuals are discarded. γ may of course depend on time on account of the evolving ecosystem.

After grid search we find x = 0.9, indicating that the random drift contributes to about 10% of survival at each cycle, and ωth = 10, in agreement with the observations prompted by the saturating behavior of ⟨ω⟩. We also find that the data cannot be approximated with a single value of γ, but they can be very well matched assuming γ = 3 for the first 5 cycles and γ = 1 for the remaining cycles, as shown in Figure 2b, green line. With the same choice of parameters, the IBEE fitness based model captures the decrease in the diversity of the DNAi ecosystem (Figure 2a, green and grey lines) and the temporal (i.e. across generations) evolution of the relative population abundance of DNAi in the three ω intervals in Figure 2, panels g, h and i.

The effect of the system size and initial conditions on the IBEE results, discussed in the Appendix 2, do not qualitatively change the models results. Error-bars on simulations have been obtained by averaging 20 independent runs, starting from the same initial conditions. As it can be observed, the variability among simulations is negligible.

The change in γ, and thus in fitness, necessary to reproduce the observed ⟨ω⟩ with the IBEE model is remarkable because it contrasts with the generally simple (fixed resource and low mutation rate) conditions of ADSE environment and indicates that survival is controlled by more than affinity to the resources, as discussed in the analysis below.

Self and mutual DNAi interactions are evolutionary drivers

While ω is certainly a key driver of the observed evolution, it is clearly not the only one. The fact that the top 10 most represented sequences come to be, in the last cycle, about 55% of the total population (Figure 2a) implies that, even among sequences with large ω, only a tiny minority come to dominate, while the largest part of them eventually disappear. Moreover, the 10 most represented sequences do not stand out for their particularly large ΔGDR or ω, as noticeable in Figure. 2c. The same data are plotted in Figure 3a as a P (ΔGDR) distribution to enable comparing the free energy distribution in the initial population (purple shading), in the final population (blue columns) and in the top 10 (black columns). These elements support the notion that survival and dominance must be also due to factors other than ω, which we thus explored.

Distribution and evolution of free energy quantifiers. a: probability distribution for the DNAi-resource binding free energy computed by NUPACK, p(⟨ΔGDR⟩), for the initial population (grey shading), for the final population (random choice of 1000 DNAi - cyan columns, top 10 most populous sequences - black columns). b: time evolution, expressed in cycles, for various mean free energies⟨ΔG⟩, normalized to their value computed on pools of random sequences. All ⟨ΔG⟩ are computed by NUPACK on sets of 1000 individuals: ΔGDR (blue dots); unimolecular self interaction ⟨ΔGself⟩ (green dots); bimolecular mutual DNAi interaction ⟨ΔGDD⟩ (red dots); mutual, self-subtracted interaction ⟨ΔΔG⟩ (yellow dots). c: scatter plot of ⟨ΔGself⟩ vs. ΔGDR computed for 1000 DNAi in the final population (blue squares). Red dots mark the point relative to the 10 most populous sequences, as identified by the labels. Note that the x-axis scale of panels a and c is the same, enabling identifying sequences. d: scatter plot computed on 104 DNAi pairs from the final population comparing ΔGselfj + ΔGself,l and ΔGDD,kl (green squares). Red dots mark the pair formed by the 10 most populous sequences, some of which identified by labels. With respect to the condition ΔGDD,kl = ⟨ ΔGself⟩,j + ⟨ ΔGself⟩,l (black line), data are on average displaced by ΔΔG ∼ 7.5 kcal/mol (yellow arrow). The online version of this article includes the following figure supplement(s) for Figure 3: Figure 3-figure supplement 1. Three examples of different relative positions for sequence-sequence attachment. Figure 3-figure supplement 2. Scheme of target-consumer interaction, from a combinatorial standpoint. Figure 3-figure supplement 3. Null model without threshold, analytical and simulated, with an even and uneven nucleotides distribution. Figure 3-figure supplement 4. Null model with/without threshold, simulated and analytical.

Figure 3b shows the value, across the 24 generations, of two other contributions to the total free energy, normalized to their value computed in random sequences, so to enable comparing their relative variations: ⟨⟨ ΔGself⟩⟩ (green symbols), the average unimolecular free energy, expressing the average strength of the internal folding of each DNAi; ⟨ΔGDD⟩ (red dots), the total bimolecular DNAi-DNAi free energy, comprising both self and mutual interactions. For comparison, we also plot the normalized value of ⟨ΔGDR⟩ (blue dots). As for the case of ΔGDR⟩, ⟨⟨ ΔGself⟩⟩ and ⟨ΔGDD⟩ are computed by randomly selecting 1000 individuals or pairs, respectively, from the population at cycle j and using the NUPACK tool to compute the values.

Figure 3b reveals that ⟨⟨ ΔGself⟩⟩ grows in time even more than ⟨ ΔGDR⟩, but with a different progression: ΔGDR grows faster in the first cycles to later saturate, while the growth of ⟨ΔGself⟩is more uniform.

Since ΔGDR is computed as the free energy of the whole DNAi-resource structure, it includes contributions of self energy associated to hairpins in the DNAi. Thus the growth of ΔGDR could actually depend on the growth in ⟨ΔGself⟩(but not the contrary). To investigate this possibility, Figure 3c displays the scatter plot between ΔGDR and ⟨ ⟨ ΔGself⟩⟩, for a randomly chosen subset of DNAi at j = 24. In evidence are the 10 points corresponding to the 10 most populous sequences. The plot shows weak or no correlation, and a relevant shift with respect to ΔGself= ΔGDR (black line), indicating that there is no dependence of ΔGDR on ⟨ΔGself⟩ and indicating that these two quantities reflect two independent driving forces in the selection mechanism of ADSE. Their behavior suggests that in the first stages of ADSE the selection is mostly dominated by the affinity with the resources, while in the later stages the requirement of a stronger unimolecular folding becomes more important.

A similar scatter plot analysis for ⟨ΔGDD⟩ yields a different outcome. Figure 3d compares ΔGDD computed for two selected DNAi (individual “k” and “l”) and the sum of ΔGself of the same two DNAi. The apparent correlation indicates that, on average, a large part of ⟨ΔGDD⟩ simply embodies the growth of self-energy, although the difference ΔΔG ≡ ΔGDD,kl − ⟨ ΔGself⟩,k − ⟨ ΔGself⟩,l (orange arrow) is non-negligible. Figure 3b shows the behavior of ΔΔG across cycles (orange squares). Despite the resulting mild growth of ΔΔG might appear not relevant, it actually indicated that ADSE, independently of ω, select strings that have higher reciprocal affinity than a random DNAi set (see Appendix 2). Indeed, as we could have obtained a significant decrease of the same quantity. It should also be noted that mutual interactions might involve the unfolding of hairpins self-structures, in which case their strength would be much larger than ΔΔG.

Self-interactions and mutual interactions can compete with the binding of DNAi to the resources either when they involve the same nucleobases or through steric hindrance. Therefore, we could expect that the increase of ΔGDR would lead to a decrease of both ⟨ ΔGself⟩ and ΔGDD. Their unfore-seen growth in the ADSE process is hence an indication of the selective advantage they convey. We interpret this behavior as an indirect sign that mutual DNAi interactions, more than just a impediment, are a major deadly threats for their survival. To avoid them, DNAi need screening. Indeed, the strong growth of self interactions, and the mild increment in mutual interactions, could represent the emergence of “defensive” strategies, as discussed below.

In our search for evolution quantifiers other than resource binding strength we also found that selection favors binding to resource sequences close to their resource 3’ terminal, away from the bead surface (see Figure 4 - figure supplement 2). While this is expected, being that terminal much less constrained and in a less crowded environment, it also provide useful clues to further analysis.

Natural history of DNAi species. a: fraction F of {DNAi} that belong to a choice of specific species as a function of the ADSE cycles, in linear (left) and logarithmic (right) scale. Arrows connect the initial condition (one individual per species) to the earliest detection via sequencing, across the 6 order of magnitudes gap (grey shading) between the population of DNAi survived at each cycle and the amount of sequenced DNAi. The same growth is assumed for species 12.1, suggesting its appearance by mutation occurred at generation 9. b-f: self interactions (b1, c1, e1), resource interactions (b2, c2, e2, d1, d2) and mutual interactions (b3, c3, e3, d1, d2, f) of selected species, sketched as per the NUPACK output. Nucleobases color coded (G -black, C - blue, A - green, T - red). Paired bases are connected. Double and single stand regions are represented as straight and curved, respectively. As in Figure 1a, terminal blocks of DNAi are marked as graphic double helices colored according to the legend of panel a, and beads as sketched yellow spheres. Yellow shading: section of DNAi complementary to resources. Pink frames: regions of hybridization between DNAi. b: interactions involving species 13.0 including its homodimerization (b3). c: interactions involving species 8.0, including its binding to 13.0 (c3). e: interactions involving species 12.1, including its homodimerization (e3). d and f: DNAi heterodimers interactions suggesting parasitism (d1), and possibly mutualism (d2) and mutual damage (f). The online version of this article includes the following figure supplement(s) for Figure 4: Figure 4-figure supplement 1. Evolution of the intra-species interaction strengths (⟨ΔGpp⟩). Figure 4-figure supplement 2. 2D probability for a consumer, shown at different experimental cycles, to have a given ⟨ω⟩ jointly with a given attachment coordinate to a target strand.

The evolution of DNAi species

While the trends in average free energies offer some insight into the nature of the ADSE ecosystem, it has also become clear that they do not enable full understanding, especially since the dominant species are not - as one could naively expect - at the extremes of the energies distributions. To gain further insight, we examined the change in numerosity of the groups of DNAi having equal sequence, which might be regarded as the “species” of ADSE. Accordingly, the 10 most populous sequences are the 10 most successful species. The evolution of the populousness, expressed as the fraction F of the total population that belong to a given species, is shown in Figure 4a for a limited set of (mainly) successful species (for a larger selection see SI). Species that are later becoming dominant “appear” in our analysis only at generation 5 in a single or a few copies. This is reasonable since we sequence a sample of ∼ 106 DNAi over the much larger population of ∼ 1012 survived individuals. Species are named after their ω value, combined with an index (e.g. “12.0”, “12.1”, “12.2” …) expressing their ranking in populousness.

ADSE species populations do not have standard patterns of evolution. This is true even when limiting the observation to the successful species in Figure 4a whose population growths are varied and do not reflect in any simple way the energy quantifiers. Conversely, species with similar energy quantifiers may have opposite fate, either becoming dominant or rapidly extinguishing or else displaying a non-monotonic population evolution. It is worth pointing out that all species start equal, each represented in the original pool by a single molecule, which on this scale would appear as a F ≈ 10−12 (black dot in Figure 4a).

We thus decided to seek further understanding on the survival of the fittest in ADSE by inspecting the “natural history” of a few species (see Tables 2, 3).

13.0 and 11.0 (orange and gray dots in Figure. 4a) are, respectively, the most and second most populous species in the last generation. It might be relevant to mention that the replica of ADSE described in the SI leads to different dominant sequences, as expected on the basis of the distinct initial pool and on the key role of randomness and sampling in the first cycles. At generation 24, 13.0 and 11.0 have grown to more than 85% and 53% of DNAi with ω = 13 and ω = 11, respectively, a further indication that ω is only one of the ingredients of the fitness. 13.0, and 11.0 likewise, have a weak self-interaction (Figure 3c), corresponding to hairpins that provide a mild defense to mutual interactions, as schematically sketched in Figure 4, panel b1. The hairpins involve most of the bases that are complementary to the resource (green shading), leaving however a few unpaired bases that might act as toehold to initiate resource binding. The resource binding takes place at its 3’ terminus (panel b2). A bit puzzled about the success of these two species we also explored their capacity of mutual interactions and found that both of them are capable of forming homodimers, as shown in panel b3 for 13.0 (binding energy marked in Figure 3d). Such dimers can bind resources at both ends (green shading regions), giving 13.0 and 11.0 self-screening and divalent binding capacity. We argue this to be a crucial drive toward success of these two species, which also explains why their growth has such an increment in the latest generations, following the probability of dimer formation.

The third most populous sequence (species 8.0) is characterized by a weak resource interaction which includes a bulge (see Figure 3c). While its survival rationale appears clear - good defensive self-binding (panel c1) and interaction targeting the 3’ end of the resource (panel c2) - it is hard to accept this could be the cause of the success. Again, inspecting the mutual interaction we found that 8.0 efficiently interacts with 13.0 (as marked Figure 3d), in a region (pink shading) that does not conflict with the capacity of 13.0 to bind to the resource (panel c3). We speculate that this parasitic capacity of 8.0 adds to the weak intrinsic binding to lead to a remarkable success of this species.

As a test to this concept, we focused on species 5.1, whose binding strength to the resource is the weakest in the top 10 species and one of the weakest among all the survivors of generation 24 (Figure 3c), a situation that cannot justify its success. However, by inspecting mutual interactions, we find its binding to 11.0 to be strong and stable (pink shading panel d1 and Figure 3d). An analogous behavior is found for species 5.0 and its interaction to 12.0 (panel d2), confirming that parasitism is an emerging successful survival solution in ADSE.

We then inspected 12.1 since it has largest ΔGDR and ⟨ ΔGself⟩ among the top 10 species (Figure 3c). Species 12.1 has all the features to succeed in ADSE: it forms a weak hairpin with the bases involved in the binding with the resources and a strong hairpin that protects the rest of the ssDNA segment (panel e1); it strongly binds to the resource with the bond ending at its 3’ terminal (panel e2); it forms homodimers capable of double resource binding, analogous to that of 13.0 (panel e3). Remarkably, 12.1 emerges only late in ADSE (Figure 4a, blue dots), a very unusual behavior when compared to the other many species we have considered. The combination of this late appearance and of the remarkable growth in F afterward suggests that 12.1 is the outcome of one of the rare mutations that can occur even with the high quality PCR that we adopted. By assuming the initial growth as with the dominant species (black arrows in panel a), we argue that such mutation occurred around generation 9, where we have found ancestors having sequence equal to 12.1 except for 4 bases.

F has a monotonic increase for most of the species that become dominant in the latest generations. This behavior is however not at all general. Non-monotonic F, growing in the first generation and decreasing to extinction afterward, is the general behavior for the majority of species that do not become extinct in the first cycle, as also shown - as mean behavior - in Figure 2h. An examples of these is species 7.X, which we plot in panel a of Figure 4 (empty dots).

More intriguing is when a non-monotonic behavior is observed for one of the dominating species. This is the case of 11.1, whose growth can be again attributed to an effective screening and a good binding to the resources, but whose decrease is harder to justify. To understand it we again investigated mutual interactions, and we found a strong 11.0-11.1 bond (sketched in panel f), which, differently from the DNAi-DNAi complex examined above, competes with resource binding. We thus speculate that the fall of 11.1 is a consequence of the rise of 11.0, which has become frequent enough to make the formation of the complex 11.0-11.1 probable, which lowers the survival probability of both species. This could also explain why species 11.0 has such an irregular growth pattern, with a slowing down when 11.1 becomes highly populated.

The insight gained by these case-studies enables appreciating how self and mutual interactions can affect the fate of species in ADSE beyond what we could discern through distributions and average quantifiers. Indeed, this analysis shows that the success of the dominant ASDE species can be achieved through different combinations of resource and mutual DNAi interactions, a fact that makes the evolution of population so varied and generation dependent and explains why, in the latest generations of ADSE, the population distribution p(ΔG) (Figure 3a) is so irregular.

Discussion and Conclusion

We have here introduced ADSE, a synthetic molecular evolution protocol that exploits sequencing technology and DNA interaction computability to provide a test-bed for key concepts in ecology and evolution. We performed experiments by adopting the most simple environment within this protocol, i.e. by holding fixed the capture sequence and minimizing mutations, to achieve a condition dominated by competition and selection, which could enable investigating the nature of fitness in this simplified evolution process.

Our experiments and their comparison with theoretical models indicate that fitness is not a simple function of the direct competitive advantage of strong binding to the resources, as it could have been naively expected. Resource binding is indeed the dominant factor in the first part of the evolution (cycles 1-5), producing a fast growth of ⟨ω⟩ compatible with a strong dependence of the survival probability on ⟨ ω ⟩(f ∝ ω3). However, as ω reaches a value corresponding to bonds of moderate stability Woodside et al. (2006) - approximately from generation 6 onward - the selective pressure related to resource binding decreases (f ∝ ω), a condition enabling to appreciate other factors at play. Which are these factors is suggested by the different growth patterns of ΔGDR and ⟨ ΔGself⟩, the former dominating the first cycles while the latter drives the later stages. However, the quantification of binding energies isn’t sufficient to predict the fate of individual species, that also depends on interaction details (location of binding and hairpin spots, secondary structures) and possibly on kinetics.

In fact, by design, DNAi can interact with themselves, both internally - forming hairpin-like structures - and mutually - leading to the formation of complexes. These interactions can either coexist-when they involve non-overlapping sequences - or they can become competitive. Our experiments reveal that, among the wide variety of conditions made available by the initial random seed, ADSE generally promotes those that combine a good interaction with the resources and a structure capable of screening from mutual DNAi interactions conflicting with binding to the beads. By exploring hybridisation-dependent DNAi secondary structures - the only ones accessible with simple analytical tools, we found three basic motifs: hairpins due to self-interactions (as for species 12.1), formation of DNAi homoduplets (as for species 13.0 and 11.0), and formation of dimers of DNAi belonging to distinct species. The latter can bring to distinct prototypical behavior: it can be the basis of parasitism, as in the case of the pairs 8.0-13.0 and 5.1-11.0, it can lead to extinction with no benefit for either species (as in the case of 15.0-12.0), or it can provide - at least in principle - mutualism. While we do not have a solid proof, clues suggest that cooperation might be the case of the interaction between species 5.0 and 12.0, since 12.0 has a weak self-defensive system and since the populations of 12.0 and 5.0 evolve very similarly (see Table 2). Finally, we cannot exclude a possible role of the formation of multi-DNAi multivalent complexes. This is the case of species 12.0 and 15.0 whose pattern of multivalent mutual interactions can, in principle, allow higher order interactions (see Appendix 2).

Fitness in ADSE is the outcome of a complex interplay of all these elements. The history of single species in Figure 4 indicates that the nature of fitness is beyond what can be captured by analyzing the probability distributions of the relevant parameters describing structure and interactions. Moreover, the variety of population evolution enlightens the fundamental fact that despite the constancy of resources, the progressive modification of the ecosystem brings about a change in the competition patterns and thus in fitness. A forthcoming work focused on a limited number of species will be devoted to better disentangle the nature of fitness in ADSE.

The evolution of the ADSE ecosystem, which corresponds to a marked decrement of its entropy - with 2/3 of the initial species become extinct in the first 5 generations - and potentially terminating with the indication of a single winner species leads instead, in the last cycles, to a significant number of dominant species all still growing at the expenses of subservient ones, indicating that in their direct competition none of them is strongly prevailing despite - or maybe thanks to - their distinct survival strategies and their very different population share. Hence, while the niche hypothesis could still be verified in the long run - beyond the experimental limitations due to PCR - it is clear that its drive is weak, suggesting that it could be overcome by environmental fluctuations, thus allowing for coexistence of different species even in a single niche environment.

Methods

Our experimental design takes advantage of a selective capture mechanism where magnetic beads carrying single-stranded DNA filaments of fixed length and sequence (resources) target DNA individuals (DNAi) (Figure 1 - figure supplement 1) present in a DNA library based on their level of complementarity. This process of selection is carried out through subsequent steps that are described in detail in the next paragraphs and represented in Figure 1-figure supplement 2.

Library design

The DNA library contains 100-nt-long sequences where a randomized central region of 50 nu-cleotides is flanked by 25-nt-long fixed sequences at both its 5’ and 3’ ends. The fixed regions provide an anchor point for primer annealing, required to perform PCR reactions during the amplification phase (see below). These terminal segments are made inactive by hybridization with oligomers of perfect complementarity (blockers), so that they are not involved in the selection phase. The blockers, as well as the fixed regions of the DNA library, have been designed to avoid any interactions with the resources. In addition, the blockers carry a phosphate group at their 3′ end to prevent them from functioning as primers during the PCR amplification. Following the above described criteria, we designed two sets of sequences (Table 1), called Oligo1 (whose results are presented in the main text) and Oligo2 that was used as a replication experiment. All the oligonucleotides used in this work were purchased from Integrated DNA Technologies, Coralville, IA, USA (Table 1).

Beads preparation

The capture of DNAi within the DNA library was performed with carboxylic acid magnetic beads (M-270 Dynabeads, Invitrogen, Carlsbad, CA) coated with the resources. The resources are 20-nucleotide-long, 5’-amino-modified oligonucleotides. Their coupling to the beads surface was performed according to the manufacturer’s instructions. Following the activation and coupling procedure, the beads were washed in Tris-HCl (50mM, pH 7.4), and stored in the same buffer in single-use aliquotes.

Sample preparation

The starting samples were prepared in 1X SSC buffer (0.15M sodium chloride, 15mM sodium citrate, pH 7.0). In detail, the Oligo1 (or Oligo2) library (0.75nmol) was mixed with the blockers (2.25nmol) in 1:3 molar ratio to saturate all the available interaction sites between the blockers and the fixed regions of the DNA library. The sample was then denatured at 95°C for 5 minutes, then slowly brought to room temperature using a thermal cycler (MasterCycler Nexus Gradient, Eppendorf, Hamburg, Germany).

Selection phase

The beads, coupled with the resources, are mixed with the sample, prepared as described above. The capture of the DNAi is carried out at 40°C for 2 hours in stirring (600rpm, ThermoMixer, Eppendorf). Once the selection phase is completed, the sample is incubated for 2 minutes on a magnet and the supernatant is removed. The beads, that are now bound to the captured DNAi, are washed three times in SSC 1X buffer to eliminate the aspecific sequences. Finally, the beads are resuspended in water and incubated for 5 minutes at 60°C to recover the DNAi from the resources. The sample is quickly placed on the magnet and after 2 minutes the supernatant containing the selected DNAi is collected.

Amplification phase

The captured DNAi are then amplified by PCR with Q5® Hot Start High-Fidelity DNA Polymerase (New England Biolabs, Ipswich, MA, USA). DNA samples were diluted 10 times and 3μl were used for each PCR reaction. The PCR was performed in a final volume of 25μL using 0.25μL of polymerase, and each primer had a final primer concentration of 2.5μM. To allow the regeneration of the single strand library, the reverse primer was designed to carry a 5′ biotin modification (see next paragraph for details). The thermal protocol was the following: i) denaturation step at 98 ? C for 2 minutes, ii) 28 cycles characterized by three thermal steps: 10 seconds at 98?C, 10 seconds at 68?C for Oligo1 (69?C for Oligo2) and 1 second at 72?C, iii) 2 minutes at 72?C. For each generation, 10 different PCR reactions were performed. The PCR products were then checked for size on a 2% TBE 1X agarose gel.

Regeneration phase

PCR products were purified by precipitation with 2.5 volumes of ethanol and 0.1 volumes of sodium acetate 2M, pH 5.2. After an overnight precipitation at -20?C, the pellet was washed with ethanol 75%, resuspended in water and quantified using the Qubit fluorometer (Invitrogen, Waltham, MA, USA). The single-strand regeneration was performed with streptavidin-coated magnetic beads (M-270 Dynabeads, Invitrogen, Carlsbad, CA) according to manufacturer’s instructions. Briefly, the PCR product was incubated for 15 minutes on a rotator with the magnetic beads. After 2 washes with binding and washing buffer (10mM Tris–HCl pH 7.5, 1mM EDTA, 0.2M NaCl) and a third one with SSC 1X buffer, an alkaline denaturation was performed by incubating the beads with 150mM NaOH for 10 minutes to induce the separation of the two DNA strands. This way it is possible to recover the unlabeled DNA strand in solution. The ssDNA is then collected and buffered with 1.25M acetic acid and TE 1X. The regenerated ssDNA is ready for the next round of selection and amplification.

Sequencing of the recovered products

To check the growth of DNAi species in our experimental model, some generations were sequenced through next-generation sequencing techniques. To prepare the sample for sequencing, the DNAi species captured as described in the “Selection phase” paragraph were first PCR-amplified using the same conditions discussed before. However, in this case, both forward and reverse primers were unlabelled. After PCR purification, performed as already described (see “regeneration phase” paragraph), the products were quantified and used to obtain libraries. About 300ng of DNA were used as starting material for the NEBNext® Ultra™ II DNA Library Prep Kit for Illumina® (New England Biolabs), and libraries were prepared following the manufacturer’s instruction. Sequencing was performed with the NextSeq™ 550 sequencer (Illumina, San Diego, CA, USA) and a paired-end strategy to obtain 75-nt long reads.

Replicas

The experimental workflow was repeated on two different sets of DNA libraries: Oligo1, whose results are described in the main text, and Oligo2, used as a replicate. The experimental conditions and procedures were the same for both sets of oligonucleotides, the only differences being the sequences of the fixed parts of the DNA libraries (and hence of the blockers and primers) and of the resources (Table 1). As a consequence, also the PCR conditions had to be adjusted, in terms of the annealing temperature that is one degree higher for Oligo2.

Eco-Evolutionary algorithm

To support experimental observation with a simple abstract model, we developed an Evolutionary Algorithm where a population of Np sequences evolves in presence of Nr < Np shorter sequences. In particular, Nr individuals in this population are selected at each cycle depending on their fitness, i.e. on their affinity to the resource, expressed via their ω with it. These survived sequences are then amplified by a factor of (roughly) Np/Nr. In this work, two types of fitness functions have been explored: the first one is merely linearly proportional to the ω of each individual (i.e., where Σω is the sum of the ω values of the whole population), while the other one is a modification of the previous fitness which sets it to be beyond a threshold value ωth. The code is written in C++, exploiting MPI Message Passing Interface Forum (2021) and Armadillo Sanderson and Curtin (2016, 2018) libraries for acceleration.

NUPACK calculations

We resorted to NUPACK for nucleotide sequences analysis, for the prediction of their free energies at equilibrium, either alone or when binding to one or two other oligomers. Schemes like those shown in Figure 4 have been obtained via the NUPACK web application Zadeh et al. (2011), while massive ΔG calculations have been performed with custom Python codes exploiting NU-PACK Python package (v4.0.0.27) Fornace et al. (2020); Dirks et al. (2007). The model is dna04 and ensemble=‘some-nupack3’; T = 40?C and [Na+] = 0.24 M, as in the experiments. Concentration of each species in the tube has been arbitrarily set to 10−6 M. Detection of hairpins and other secondary structures has been performed visually (thanks to the oxView software Bohlin et al. (2022); Poppleton et al. (2020)).

Supporting information

eLife Assessment

In this important study, the authors develop a promising experimental approach to a central question in ecology: What are the contributions of resource use and interactions in the shaping of an ecosystem? For this, they develop a synthetic ecosystem set-up, a variant of SELEX that allows very detailed control over ecological variables. The evidence is convincing, and the work should be of broad interest to the ecology community, leading to further quantitative studies.

Reviewer #1 (Public Review):

This work describes a new and powerful approach to a central question in ecology: what are the relative contributions of resource utilisation vs interactions between individuals in the shaping of an ecosystem? This approach relies on a very original quantitative experimental set-up whose power lies in its simplicity, allowing an exceptional level of control over ecological parameters and of measurement accuracy.

In this experimental system, the shared resource corresponds to 10^12 copies of a fixed single-stranded target DNA molecule to which 10^15 random single-stranded DNA molecules (the individuals populating the ecosystem) can bind. The binding process is cycled, with a 1000x-PCR amplification step between successive binding steps. The composition of the population is monitored via high-throughput DNA sequencing. Sequence data analysis describes the change in population diversity over cycles. The results are interpreted using estimated binding interactions of individuals with the target resource, as well as estimated binding interactions between individuals and also self-interactions (that can all be directly predicted as they correspond to DNA-DNA interactions). A simple model provides a framework to account for ecosystem dynamics over cycles. Finally, the trajectory of some individuals with high frequency in late cycles is traced back to the earliest cycles at which they are detected by sequencing. Their propensities to bind the resource, to form hairpins, or to form homodimers suggest how different interaction modes shape the composition of the population over cycles.

The authors report a shift from selection for binding to the resource to interactions between individuals and self-interactions over the course of cycles as the main drivers of their ecosystem. The outcome of the experiment is far from trivial as the individual-resource binding energy initially determines the relative enrichment of individuals, and then seems to saturate. The richness of the population dynamics observed with this simple system is thus comparable to that found in some natural ecosystems. The findings obtained with this new approach will likely guide the exploration of natural ecosystems in which parameters and observables are much less accessible.

My review focuses mainly on the experimental aspects of this work given my own expertise. The introduction exposes very convincingly the scientific context of this work, justifying the need for such an approach to address questions pertaining to ecology. The manuscript describes very clearly and rigorously the experimental set-up. The main strengths of this work are (i) the outstanding originality of the experimental approach and (ii) its simplicity. With this setup, central questions in ecology can be addressed in a quantitative manner, including the possibility of running trajectories in parallel to generalize the findings, as reported here. Technical aspects have been carefully implemented, from the design of random individuals bearing flanking regions for PCR amplification, binding selection and (low error) amplification protocols, and sequencing read-out whose depth is sufficient to capture the relevant dynamics. One missing aspect in the data analysis is the quantification of the effect of PCR amplification steps in shaping the ecosystem (to be modeled if significant). In addition, as it stands the current work does not fully harness the power of the approach. For instance, with this setup, one can tune the relative contributions of binding selection vs amplification for instance (to disentangle forces that shape the ecosystem). One can also run cycles with new DNA individuals, designed with arbitrarily chosen resource binding vs self-binding, that are predicted to dominate depending on chosen ecological parameters.

Reviewer #2 (Public Review):

Summary:

In this manuscript, the authors introduced ADSE, a SELEX-based protocol to explore the mechanism of emergency of species. They used DNA hybridization (to the bait pool, "resources") as the driving force for selection and quantitatively investigated the factors that may contribute to the survival during generation evolution (progress of SELEX cycle), revealing that besides individual-resource binding, the inter- and intra-individual interactions were also important features along with mutualism and parasitism.

Strengths:

The design of using pure biochemical affinity assay to study eco-evolution is interesting, providing an important viewpoint to partly explain the molecular mechanism of evolution.

Weaknesses:

Though the evidence of the study is somewhat convincing, some aspects still need to be improved, mostly technical issues.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1:

This work describes a new and powerful approach to a central question in ecology: what are the relative contributions of resource utilisation vs interactions between individuals in the shaping of an ecosystem? This approach relies on a very original quantitative experimental set-up whose power lies in its simplicity, allowing an exceptional level of control over ecological parameters and of measurement accuracy.

In this experimental system, the shared resource corresponds to 10^12 copies of a fixed single-stranded target DNA molecule to which 10^15 random single-stranded DNA molecules (the individuals populating the ecosystem) can bind. The binding process is cycled, with a 1000x-PCR amplification step between successive binding steps. The composition of the population is monitored via high-throughput DNA sequencing. Sequence data analysis describes the change in population diversity over cycles. The results are interpreted using estimated binding interactions of individuals with the target resource, as well as estimated binding interactions between individuals and also self-interactions (that can all be directly predicted as they correspond to DNA-DNA interactions). A simple model provides a framework to account for ecosystem dynamics over cycles. Finally, the trajectory of some individuals with high frequency in late cycles is traced back to the earliest cycles at which they are detected by sequencing. Their propensities to bind the resource, to form hairpins, or to form homodimers suggest how different interaction modes shape the composition of the population over cycles.

The authors report a shift from selection for binding to the resource to interactions between individuals and self-interactions over the course of cycles as the main drivers of their ecosystem. The outcome of the experiment is far from trivial as the individual resource binding energy initially determines the relative enrichment of individuals, and then seems to saturate. The richness of the population dynamics observed with this simple system is thus comparable to that found in some natural ecosystems. The findings obtained with this new approach will likely guide the exploration of natural ecosystems in which parameters and observables are much less accessible.

My review focuses mainly on the experimental aspects of this work given my own expertise. The introduction exposes very convincingly the scientific context of this work, justifying the need for such an approach to address questions pertaining to ecology. The manuscript describes very clearly and rigorously the experimental setup. The main strengths of this work are (i) the outstanding originality of the experimental approach and (ii) its simplicity. With this setup, central questions in ecology can be addressed in a quantitative manner, including the possibility of running trajectories in parallel to generalize the findings, as reported here. Technical aspects have been carefully implemented, from the design of random individuals bearing flanking regions for PCR amplification, binding selection and (low error) amplification protocols, and sequencing read-out whose depth is sufficient to capture the relevant dynamics.

:

We thank the reviewer for summarizing our work and the main findings in a very clear and effective manner.

One missing aspect in the data analysis is the quantification of the effect of PCR amplification steps in shaping the ecosystem (to be modeled if significant). In addition, as it stands the current work does not fully harness the power of the approach. For instance, with this setup, one can tune the relative contributions of binding selection vs amplification for instance (to disentangle forces that shape the ecosystem). One can also run cycles with new DNA individuals, designed with arbitrarily chosen resource binding vs self-binding, that are predicted to dominate depending on chosen ecological parameters. I have three main recommendations to the authors:

1. PCR amplification steps (and not only binding selection steps) should be taken into account when interpreting the outcome of experiments.

1. More generally, a systematic analysis of the possible modes of propagation of a DNA molecule from one cycle to the next, including those considered as experimental noise, would help with interpreting the results.

1. Testing experimentally the predictions from the analysis and the modelling of results would strengthen the case for this approach.

Despite its conceptual simplicity, our approach has indeed a few experimental handles that enable exploring a relevant variety of conditions much beyond those described in this paper, of which we are very aware. These involve selection vs. amplification or set the stage to explore competition, parasitism or cooperation among specific species, as the reviewer points out, but also introduce mutations and explore the kinetics of evolution in static or dynamic environments. Ongoing experiments are considering some of these conditions. We modified the text to mention more explicitly these possibilities, which are now mentioned in p11 lines 376-378 and lines 416-417. The three points raised by the reviewer helped us to further improve and clarify strengths and limitations of our work, as detailed below.

Regarding the first point, here are my suggestions :

Run one cycle of just amplification vs 'binding + amplification', or simply increase the number of PCR cycles (and subsample the product) to check whether it impacts the population composition, in particular for sequences with predictions derived from the current analysis.

The point raised by the reviewer is indeed very relevant and not discussed in our manuscript. Prompted by the reviewer’s comment, we performed two new experiments to distinguish resource-binding selection from PCR amplification effects.

First, we performed a negative control experiment in which we performed the “selection step” with bear beads, i.e. beads without with no DNA grafted on them. We then compared the results with the corresponding results of the original experiments on Oligo 1 and 2.

After 6 cycles, the most abundant sequence in the negative dataset has a relative occurrence of 0.05%, whereas the dominant strand in Oligo 1 and Oligo 2 has an abundance of 8% and 16%, respectively, i.e. 40-80 times larger.

This indicates that the drift due to non-specific binding + PCR amplification is at least two orders of magnitude smaller than the selection induced by the affinity with the resource.

This results are now cited in p14 lines 468-470, and described in Appendix 1, Experimental controls.

Second, we tested the effect of PCR amplification on the selection process. We exploited the fact that we have aliquots for each generation of our evolution experiment, which we sampled and saved after PCR and before sequencing. We thus chose a specific generation - specifically generation 9 from Oligo 1 experiments - and performed another PCR round we proceeded directly to sequencing with no beadsselection step. We then compared the ensemble of oligos obtained in this way, which we named Oligo 1 “cycle 9 replica”, with both the original Oligo1 cycle 9, and with Oligo1 cycle 10.

We sampled 20 times 4 x 10^5 sequences from the cycle 9 dataset, from cycle 9 replica and from cycle 10 with a bootstrap approach. To compare the three systems we extracted the fraction of the population of each covered by the 10 most abundant individuals. The results are shown in Figure 2 - Figure Supplement 4. In the figure caption further details on the analysis can be found. The similarity between cycle 9 and cycle 9 replica and the marked difference between cycle 9 replica and cycle 10

indicates that the relevant part of the selection is indeed performed by the resourcebinding mechanism, while drifts induced by PCR play a secondary role.

As a further check, we compared the specific sequences across the 20 samples in cycle 9 and cycle 9 replica datasets and found that the 10 most abundant sequences are almost always the same. In particular, the first 8/9 are always the same, possibly shuffled.

These new pieces of evidence are now cited in p14 lines 483-484 and described in Appendix 1, Experimental controls.

Sequencing read-out includes the same PCR protocol as the one used for amplification steps, so read-out potentially has an effect on the composition of the ecosystem. Again, varying the number of PCR cycles is a direct way to test this.

The PCR amplification involved in the read-out might have a minor effect on the sequencing outcome but not on the composition of the ecosystem. In fact, the sample that undergoes sequencing is taken from the pool at each cycle, and not inserted back into it. Thus, it does not participate in the following selection steps. This is specified in the text at p3 line 104

Could self-interactions (hairpins of homodimers) benefit individuals during amplification steps? The role of self-interactions during binding selection steps could also be tested directly over one cycle (again varying the relative weight of the binding vs amplification to disentangle both).

Our choice of conditions for PCR amplification were thought to minimize effects of this type. PCR amplification is carried out at 68 C, a temperature at which, given the level of self and mutual complementarity in the sequences analyzed in the text, hairpins or homodimers should be melted and thus have no effect. This is specified in the text at p. 14 lines 479-480 However, if an effect is present, it gives a disadvantage (rather than an advantage) to self-interacting individuals. For the amplification step we used Q5® Hot Start HighFidelity DNA Polymerase, which does not possess strand displacement activity. Therefore, in theory, if during amplification the polymerase encounters a double strand portion, it stops and synthesizes only a truncated product, which will be then lost during the purification step. In other words, sequences with secondary and/or tertiary structures are less likely to be amplified during the polymerization step. As a consequence, a DNAi that is characterized by this kind of structures, will be negatively selected even in the case of optimal binding to the resource, and will be underrepresented in the pool.

About the second point:

Regarding the effect of sampling (sequencing read-out), PCR amplification errors: explicitly check the consistency of observations with the expected outcome, in the methods section (right now these aspects are only briefly mentioned in the main text), which would highlight again the level of control and accuracy of the system.

Hoping to have well interpreted the request, we performed a technical replicate sequencing Oligo 1 cycle 9 again and analyzed the sequences that have at least 100 reads (corresponding to 27.42% of the total reads). We find that among the 800 DNA species that have at least 100 reads, 93.6% are found in both replicates. All the nonoverlapping sequences have very low abundance, close to 100.

Moreover, we compare the population size of each DNA species between the two replicas, after having equalized the database sizes. The results are now cited in p14 lines 509-510, In Appendix 1, Experimental Controls and shown in Figure 2-figure supplement 3, where we plot the ratio of the number of reads in the two replicates for each sequence as a function of the number of reads in one. We found an average of 0.965 with a standard deviation of 0.119. High fluctuations are found in the most rare species, as expected.

We think this evaluation indeed strengthens the solidity of our results.

I have a small concern about target resource accessibility: is there any spacer between the ssDNA and the bead? The methods section does not mention any, and I would expect such a proximity between the target DNA and the bead to yield steric repulsion that impedes interactions with random DNA individuals.

Yes, there is a 12-carbon spacer between the bead and the resource, which was inserted exactly to make the resource more accessible. This information is now available in Table 1 of Supplementary Information detailing the sequences used in the experiment. However, as now described in the text (p8 lines 284-286), we observe that the interaction with the resource is always shifted to the 3', the terminal furthest from the bead, indicating some residual issue of accessibility to the resource sections closest to the bead.

Regardless of the existence of a spacer, binding of random DNA molecules to beads instead of the target DNA constitutes a potential source of noise (described for now as '1-x' in the IBEE model), which can be probed by swapping targets, selecting without target etc.

This issue is addressed by the test with bare beads described above, in which we found little effects, corresponding to small 1−𝑥 value.

Is there any recombination potentially occurring during amplification steps? This could be tested with a set of known molecules amplified over 24 amplification steps in a row (no binding step).

It is possible for recombination to occur during the amplification steps. In Appendix 2, the section "By-Product Formation from PCR Amplification", discusses PCR byproducts as aberrant forms of amplification, such as recombination events. We adopted several strategies to limit by-product formation, such as: i) use of “blockers” characterized by a phosphate group at 3’ end (thus inhibiting their usage during the amplification and allowing a better control of the reaction conditions over the PCR cycles), ii) a high annealing temperature (to limit the possibility of a spurious primer annealing to the random region), iii) fewer PCR cycles, iv) a high primer concentration, v) a very short elongation step (all these strategies have been implemented to avoid a possible mispriming event between different DNAi, and the formation of concatemers). However, the formation of by-products is a problem inherent to the technique: in fact, it is a known issue for classical SELEX technology (Tolle et al. 2014), mainly due to the random region within the DNAi. Q5® Hot Start High-Fidelity DNA Polymerase (New England Biolabs, Ipswich, MA, USA) has an error rate of <0.44 x 10-6/base.

In classic SELEX technology, the average number of selection cycles is 10. This limitation is partly due to the increase in PCR by-products. As we can see from Figure 2 Supplementary Figure 1, the percentage of PCR by-products is less than 20% at cycle 12, and then increases dramatically in the following cycles. We are performing a series of experiments with known and limited sequences to verify and better understand the phenomenon for future applications of the SEDES platform. On this issue we decided not to modify the manuscript since we think it is already well discussed in Appendix 1.

And the third point:

Perform one cycle (or a few cycles) with random DNA individuals, the most frequent individuals at the end of the current experiment, newly designed individuals with higher binding affinity to the target than currently dominating individuals, newly designed individuals with higher propensity to form hairpins or to form homodimers. Such experimental testing of predictions from the data analysis/modeling, typical of a physics approach, would illustrate the level of understanding one can reach with a simple yet powerful experimental setup.

We perfectly agree that the approach we propose and the set of results we obtained call for further investigations that could strengthen analysis and modeling. The final aim we envisage is the understanding, within this simplified approach, of key evolutionary factors such as fitness. Indeed, becoming able to write an explicit fitness function would be a significant new contribution to the understanding of evolutionary processes, even within the limited settings of the ADSE approach, as discussed in the conclusions of the manuscript.

However, undergoing such an analysis is a long and expensive job, which we have started and will be completed in a not immediate future. For this reason, given the already significant body of results we are presenting here, we prefer to keep this paper confined to the study of the evolution of a random DNAi population and discuss in a future contribution the behavior of smaller designed sets of competing, collaborating or parasitic individuals.

Looking ahead, additional stages of investigations will also include mutations - to investigate the kinetics of speciation, and, in an even further stage, the interplay between evolution kinetics and dynamical mutation of resources.

I have a few smaller points:

It would be very useful to provide the expected dynamic range of binding free energies (in terms of DeltaG and omega): what is the maximum binding free energy for the perfect complement?

The NUPACK-computed binding free energy of a 20 basis-long oligomer complementary to the resource (𝜔=20) is -24.36 Kcal/mol for Oligo1 and -23.08 Kcal/mol for oligo 2. This is the best answer we can offer to the reviewer’s request, since the maximum binding free energy of DNAi individuals (much longer than the target strand) would include contributions from the unpaired bases. Indeed, the values give above are approached by the left tail of the distribution of Fig. 3a, which however includes DNAi self-energies.

The perfect complement binding free energy is now cited in the text as a reference for the dynamical range of DeltaG (p4 lines 151-152).

How is the number of captured DNA molecules quantified? Is 10^12 measured, estimated, or hypothesized?

The number of sequences was calculated from data obtained from 260 nm absorbance quantification. We have now added this information in the Methods, Selection Phase” section.

Reviewer #2:

Summary:

In this manuscript, the authors introduced ADSE, a SELEX-based protocol to explore the mechanism of emergency of species. They used DNA hybridization (to the bait pool, "resources") as the driving force for selection and quantitatively investigated the factors that may contribute to the survival during generation evolution (progress of SELEX cycle), revealing that besides individual-resource binding, the inter- and intra-individual interactions were also important features along with mutualism and parasitism.

Strengths:

The design of using pure biochemical affinity assay to study eco-evolution is interesting, providing an important viewpoint to partly explain the molecular mechanism of evolution.

Weaknesses:

Though the evidence of the study is somewhat convincing, some aspects still need to be improved, mostly technical issues.

Major:

1. There are a few technical issues that the authors should clarify in the manuscript to make the analysis more transparent:

1.1) To my understanding, it is difficult to guarantee the even distribution of different species (individuals) in the initial individual pool. Even though the authors have shown in Fig. 2a that the top 10 sequences take up ~ 0% in the pool, it remains unclear how abundant these top and bottom representative sequences are, given the huge number of the pool (10E15). Can the author show the absolute number of these sequences in different quantiles? Please show both Oligo sets.

: First, we thank the reviewer for both positive and critical comments that have guided us in reformulating or clarifying some messages of our work.

As for this specific point: 10E15 is a small number compared to 4^50 = 10E30, the number of possible sequences of length 30. Thus, we don’t expect more than one individual per sequence in the initial pool. However, sequencing requires a preparation amplification, which may lead to detecting a few sequences with more than one individual.

Specifically, in the initial pool of Oligo 1, the most abundant individual (of sequence GAACTAAAGGGGCGGTGTCCACTTGCCTGTAGTGGTTATCAGTCCGGTTG)has 3 copies. The 0.7% of the sequences has 2 copies, while the vast majority of strings (99.3% on a sample of about 1.5 x 10E6 sequenced DNAi) is present in one copy only. A similar situation holds for Oligo 2, with 4 DNAi present in 3 copies and the 0.8% of the sequences (in a pool of 2 x 10E6 DNAi) in 2 copies.

It is worth noticing that none of the 10 most abundant species in the last cycle is present in the sample. Indeed, the fraction of the pool which is sequenced is removed from the population that undergoes evolution (as now specified in p2, line 104). We specified in the text (p2, lines 69-70, p3 lines 94-96) the fact that in the initial pool no sequence is expected to be present in more than one individual.

1.2) The author claimed that they used two different oligo sets (Oligo1 and Oligo2) in this study. It is unclear which data was used in the presentation. How reproducible are they? Similar to this concern, how reproducible if the same oligo set was used to repeat the experiment?

The oligo used in the main text was declared in Methods, Replica section. It is now declared also in the main text (p3 lines 106-108 and in the captions of Figure 2, Figure 3 and Figure 4). Reproducibility is addressed in: Figure 2-figure supplement 5; Figure 2-figure supplement 6; Appendix 2: Results of the experimental replica.

It should also be noted that two starting pools of random 50mers are necessarily disjoint sets for the same reason discussed in the previous answer: the probability of common sequences in two 10E15 selections from a 10E30 is negligibly small. Thus, it is expected that each time a new evolution experiment is started, different dominant sequences are found. However, the statistical properties of the DNAi pool during the evolution process of Oligo1 and Oligo2 are similar as discussed in Appendix 2 of the paper.

1.3) PCR and illumina sequencing itself introduced selection bias. How would the analysis eliminate them? The authors only discussed the errors created during PCR cycles (page 3, lines 115-122). However the PCR itself would prefer to amplify some sequences over the others (e.g. with high GC content). Similarly, the illumina sequencing would be difficult to sequence the low complexity sequences. How would this be circumvented?

Yes, both PCR and Illumina sequencing have some known biases in the amplification process (e.g. sequencing of homopolymers or amplification of GC-rich sequences) that are intrinsic to the used techniques. Regarding PCR, we implemented a thermal protocol optimized for our chosen experimental setup, characterized by very short denaturation, annealing and amplification steps performed at high temperatures. Regarding Illumina sequencing, we can’t rule out a bias against specific sequences (e.g, homopolymers), which however should not be captured during the selection step, due to the design of the resource. Also, the libraries subjected to sequencing are characterized by a low complexity: according to the experimental design, the first and last 25 nucleotides are the same for all DNAi, the only differences being in the central 50 nt-long sequence. It is known that a low complexity library might encounter problems during sequencing due to the design of Illumina instruments: nucleotide diversity, especially in the first sequencing cycles, is critical for cluster filtering, optimal run performance and high-quality data generation. To overcome this limitation, the obtained libraries were run together with more complex and diverse library preparations: the ADSE sequences were about 1-2% of the total reads per run, corresponding to only a few million reads.

This discussion is now in Appendix 1, Intrinsic limitations of the molecular approach.

1.4) Some DNA sequences would bind to the beads instead of the resource sequence coated on them. Should the author run the experiment using bead alone as a control?

: We performed a negative control experiment in which we performed the “selection step” with bear beads, i.e. beads without with no DNA grafted on them. We then compared the results with the corresponding results of the original experiments on Oligo 1 and 2.

After 6 cycles, the most abundant sequence in the negative dataset has a relative occurrence of 0.05%, whereas the dominant strand in Oligo 1 and Oligo 2 has an abundance of 8% and 16%, respectively, i.e. 40-80 times larger.

This indicates that the drift due to non-specific binding (+ PCR amplification) is at least two orders of magnitude smaller than the selection induced by the affinity with the resource.

This part is now discussed in Appendix 1, Experimental controls.

1. It would be interesting to study the impact of environmental factors, for example, changing pH, salt concentration, and detergent. Would these factors accelerate/decelerate the evolution?

We agree that the approach we propose and the set of results we obtained call for further investigations. However, performing these additional experiments, which would require a minimum of 6 generations each, is a long and expensive job, which we have started and will not be completed in the near future. For this reason, given the already significant body of results we are presenting here, we prefer to keep this paper confined to the study of the evolution of a random DNAi population in the selected conditions and leave the exploration of new conditions, potentially opening new evolutionary scenarios, to a future contribution. In fact, our aim was to show that through our platform we can indeed observe fundamental elements of evolution in a non-biological system, which, in the set of chosen parameters, we do.

1. The concentration of individual oligo is apparently one of the most important factors in determining the interactions. In later cycles, some oligos become dominant, namely with extremely higher concentrations compared to their concentration in earlier cycles. This would definitely affect its interaction with resources, or self-interaction, or interaction with other oligos in the pool. However, the authors failed to discuss this factor, which may explain the exponential enrichment in later cycles.

We agree with the reviewer that this is an important point, but we disagree that we have not discussed it. We introduce the topic at the end of the “Null Model and Eco-evolutionary Algorithm”, where we comment on the change of the gamma parameter by saying that there must be a shift in the evolution process, first dominated by the interactions with the resources, and in later stages by some other factors (lines 227230) that we then discuss in “Self and mutual DNAi interactions are evolutionary drivers”. In this latter chapter and in the following, we indeed discussed the effects of mutual and self interactions between DNAi.

Indeed, a key point in our paper is the change in the gamma parameter necessary to match the IBEE model to experiments, as it is now more openly stated (p5 lines 217218 where we also mention figure 2-supplement 8 which clearly shows the necessity of a variable gamma). The two regimes enlightened by the gamma value must reflect a change in the competition for the resources and interactions among species. In the first generations, where the diversity of species is large (there are few strings for each species) and binding to the resources generally very week (small ), the affinity with the resource is the main driving force (fast growth of ), while mutual interactions remain too random to favor any species in particular. In the later cycles instead, when becomes large enough to provide a significant stability to the resource-binding of the majority of species, the dominating species compete more intensively on the basis of their structure and capacity of self-defense, parasitism and mutualism, a condition in which evolution affects more modifications in sequences than in .

Certainly, our understanding of this shift is based on statistical behavior and it is inferential, based on the study of specific DNAi described in the last part of the manuscript. For a better molecular model, more experiments with selected DNAi competing, cooperating or being parasitic would be necessary, with the final aim of defining a predictive fitness function. Alas, this requires months of further investigation. :

1. The author observed the different behaviors of medium 𝜔 in early and late cycles, referring to Fig 2h. Using the IBEE model, they found out it is the change of gamma. However, the authors did not further discuss the molecular mechanism. It could be very interesting to understand the evolutionary change of these individuals.

This comment might be related to the previous one. It is true that our discussion and understanding of the whole process is statistical, and misses a molecular model to predict the value of gamma.

However, the specific behavior that the reviewer asks about (those in Fig. 2h) is not related to the change in gamma. Even if gamma remains as in the first part of the evolution (gamma = 3), the species with overlap between 6 and 10 would first grow in number and later decrease. Indeed, during the first cycles they have an advantage with respect to the majority of species with lower maximum overlap, a condition that favors their amplification. However, in the second stage of the evolution dominant species with a larger affinity emerge and outcompete the individuals of this class. We added a sentence in the text to clarify this point (p7 lines 227-229).

1. In Figure 2f, some high w become quite missing. Should the authors give some interpretation? It is not observed in cycle 12 though (panel e).

Such an effect is just due to under-sampling. In a pool of 10^n oligomers, any sequence with a given 𝜔 with P(omega) < 10E-n will have a vanishing probability to appear in that sample.

At cycle 12 the overall number of sequenced strands is larger than at cycle 24, due to the growing presence of PCR by-products. Thus, the right tail of the cyan distribution at the last cycle is sampled with less accuracy than at cycle 12. We have added a sentence in the revised manuscript (p5 lines 177-178) to clarify this point.

1. It would be interesting to further explore if another type of selection resource is used, for example protein that binds to particular sequences, i.e. transcription factors. Previous studies have used a large amount of sequence-specific transcription factors to run SELELX. Since the data have existed there, why not explore?

This is an interesting suggestion: can we use data from “ordinary” SELEX favoring specific sequences to explore sequence evolution? Two limitations make us a bit skeptical on this path: first, the consensus sequences of DNA-binding proteins are rather short and typically target dsDNA rather than ssDNA; second, the free energy of interaction is known only for the consensus sequence but not for sequences with all possible mutations with respect to the consensus sequence, making very hard to develop any molecular understanding of the process.

Minor:

1. There is no figure legend or in-text citation of Figure 2b.

1. Please correct "⁃C" with "{degree sign}C" in lines 470, 471, 472, 477 et al.

1. Typos and grammar issues should be corrected. Examples are shown below (but not limited to these only):

mixed use of past and present tense.

Line 152, "basis" should be "bases".

Line 277, "a impediment" should be "an impediment"

Line 278, "a major deadly threats" should be "major deadly threats"

:

We are sorry for the mistakes, and we have corrected them. Many thanks to the reviewer!

Version 2

Abstract

Abstract

The understanding of eco-evolutionary dynamics, and in particular the mechanism of coexistence of species, is still fragmentary and in need of test bench model systems. To this aim we developed a variant of SELEX in-vitro selection to study the evolution of a population of ∼ 1015 single-strand DNA oligonucleotide ‘individuals’. We begin with a seed of random sequences which we select via affinity capture from ∼ 1012 DNA oligomers of fixed sequence (‘resources’) over which they compete. At each cycle (‘generation’), the ecosystem is replenished via PCR amplification of survivors. Massive parallel sequencing indicates that across generations the variety of sequences (‘species’) drastically decreases, while some of them become populous and dominate the ecosystem. The simplicity of our approach, in which survival is granted by hybridization, enables a quantitative investigation of fitness through a statistical analysis of binding energies. We find that the strength of individual-resource binding dominates the selection in the first generations, while inter and intra-individual interactions become important in later stages, in parallel with the emergence of prototypical forms of mutualism and parasitism.

Introduction

A central effort in theoretical biology and ecology is to provide an effective description of the intimate, but often subtle, relationship between a given environment and the evolution of its ecosystem. In the case of simple environments without geographical isolation and physical barriers, as the one here considered, the emergence of species and phenotyping clustering Davis (1943); Rundle and Nosil (2005); Keymer et al. (2012); Gupta et al. (2021) is generally considered an outcome of the competition for the limited available resources Dieckmann and Doebeli (1999); Pigolotti et al. (2007); de Aguiar et al. (2009); Anceschi et al. (2018). The coexistence of stable species is ecologically understood in terms of “niches”, indicating the unique role and position that a particular species occupies within an ecosystem. According to the “niche hypothesis” Chase and Leibold (2009); Peterson (2011); Anceschi et al. (2018), biodiversity is limited by the number of “niches” (or types of resources) that are present since no two species can occupy the same niche indefinitely, as one would eventually out-compete the other Levin (1970); Gupta et al. (2021).

Fitness, which quantifies species reproductive success and thus also their relationship with the environment De Visser and Krug (2014), cannot generally be defined in a predictive way even in the most idealized systems Thurner et al. (2010); Wiser and Lenski (2015). Indeed, despite the efforts to identify simple case study systems, the evolution of populations formed by a variety of species remains difficult to model and to quantitatively characterize because of the inherent complexity of ecosystems and living beings, of the large number of potentially relevant variables of difficult access, and of the role of stochasticity Catalán et al. (2017).

In this scenario, introducing new tools to explore and test eco-evolutionary models, concepts and interpretations appear as the best strategy toward new understanding Solé (2016). Along this line, various synthetic biological platforms have been proposed in the last years Ichihashi et al. (2013); Tizei et al. (2016); Parrilla-Gutierrez et al. (2017); Kauffman et al. (2018); Adamala and Szostak (2013); Katla et al. (2023) that exploit different principles and mechanisms, and focused on in-vivo, in-vitro, ex-vivo, or in-silico approaches. For example, in Ichihashi et al. (2013) a “cell-like” model system is proposed, in which the evolution of one long genomic RNA (> 2,000 nt-long) was investigated in detail under the action of a selective pressure ultimately provided by its biological meaning. Despite the tremendous simplification provided by this approach, the constructed artificial cell is still a complex system that includes ribosomes, lipids, translation factors, accessory proteins, tRNAs and amino acids. While most of the synthetic biological platforms are based on the establishment of cell-like, compartmentalized systems involving complex biomolecular milieus and processes Tizei et al. (2016), a few are devoted to investigate, with distinct strategies, the competition between two coexisting vescicle-based species competing for resources (vescicle-forming molecules) in a non-biochemical context Adamala and Szostak (2013); Katla et al. (2023). These articles demonstrate that typical evolution concepts, such as competition or resources and niche exclusion principle, can be applied to “non-biological” systems.

We here propose a synthetic eco-evolutionary scheme, with no cell-like compartmentalization and no connection to the molecular biology of the cell: no coding sequences, no translation, no proteins involved. We consider an ecosystem formed by a large crowd (∼ 1015) of distinct molecular individuals interacting with each other and competing for survival in an environment with fixed resources, and we focus on the process by which selection and competition drive the emergence of dominating species. Specifically, we study the evolution of a pool of 50-base-long single strand DNA oligomers with random sequences, a choice that ensures that in the initial solution each molecule is unique. The mechanisms of survival and mutual interactions are based on DNA hybridization. By exploiting the wealth of tools and knowledge about DNA code selective pairing and DNA synthesis, amplification and sequencing, we create a condition in which a limited number of accessible and computable variables controls the destiny of an ecosystem formed by biological molecules that evolve in a non-biological way, i.e. with no reference to the biological meaning of their sequence.

The results of our work, demonstrating the intimate connection between fitness and ecological interactions, belong to the growing body of studies Fussmann et al. (2007); Vetsigian (2017); Camacho Mateu et al. (2021) showing that ecological and evolutionary processes are strictly related in the emergence and maintenance of species.

Results

Affinity-based DNA Synthetic Evolution

We introduce here a variant of SELEX for in-vitro synthetic evolution of oligonucleotides to develop protein-binding aptamers Ellington and Szostak (1990); Tuerk and Gold (1990). In standard SELEX protocol, the evolving oligonucleotides are selected at each cycle by their interactions of with the target protein. In the experiment here reported, we implemented a selective mechanism based on the affinity capture provided by magnetic beads carrying single-stranded DNA (ssDNA) filaments of fixed length L = 20 and sequence, that act as targets (or resources, Figure 1a). Selection is thus primarily based on the sequence of the DNA individuals and its level of complementarity to the targets. This marks a significant difference with SELEX, in which the aptamer-protein interaction depends instead on higher order factors such as the secondary structure of the oligonucleotides and the variety of binding sites on the folded protein. Being these hard to model, predict and control, SELEX has never been considered, to the best of our knowledge, as a useful experimental test bench to understand evolution.

Affinity-based DNA Synthetic Evolution (ADSE). a: structure of the DNA oligomers participating in ADSE as individuals (DNAi) and target resources. b: steps in the ADSE. The process starts with a random-sequence DNAi population. The capture by magnetic bead-conjugated resources provides the selection: bead-bound DNAi are amplified to form the new generation, a small fraction of which is sequenced by massive parallel sequencing. The rest of the original solution is discarded. Red arrows mark the steps of each ADSE cycle. c: possible interaction motifs involving DNAi.

The online version of this article includes the following figure supplements:

Figure 1-figure supplement 1 Detailed structure of DNA individuals.

Figure 1-figure supplement 2 Detailed scheme of the Affinity-based DNA Synthetic Evolution protocol.

In our Affinity-based DNA Synthetic Evolution (ADSE) protocol, evolution starts from an initial pool of DNA Individuals (DNAi), chosen to be of fixed length L = 50 and random sequence. Each sequence indicates a ‘species’. Since the potential molecular variety is 450 ∼ 1030 while our experiments use about 1015 initial molecules, each species in the initial pool has only one DNAi. The following evolution process is sketched in Figure 1b, Figure 1-figure supplement 2, and is given by three steps.

Selection: the seed population is mixed with a given amount of dispersed capture beads. After a suitable incubation time, the beads are extracted from the solution and the bound oligomers released and saved. The rest of the original solution is discarded;

Amplification: the pool of “survived” oligomers is PCR amplified about 1000 times to recover the initial molarity;

Sequencing: a small portion of the amplified sample (1-3 106 molecules) is analyzed by massive parallel sequencing (MPS). These molecules are thus removed from the evolving pool.

These steps constitute a cycle - one generation of evolution - that we repeated up to 24 times in two independent evolution histories, which we refer to as “Oligo 1” and “Oligo 2”. In the following, we present in detail the results of the Oligo 1 evolution, while Oligo 2 results are described in the figure supplements.

In actuality, to enable amplification and sequencing, DNAi are built by flanking the 50mer with two 25 base long fixed sequences that enable primer binding for a total length of 100 bases. Such two terminal segments of the DNAi are made inactive during selection by hybridization with oligomers of perfect complementarity, as sketched in Figure 1a, Figure 1-figure supplement 1.

A key feature of ADSE is that DNAi can interact not only with the resources, but also with itself and with each other, and also potentially form complexes binding to the affinity beads as summarized in Figure 1c. Indeed, the choice of a length of 50 for the DNAi interaction was thought to enable - in principle - simultaneous resource and mutual binding.

PCR amplification can be of high fidelity or error-prone, the latter choice enabling genetic drift and, potentially, speciation. Since our primary goal was to first investigate fitness in a pool of competing species within a given niche, defined in our case by the 20 base long resources, we opted for a high-fidelity amplification, and left the investigation of high mutations regimes to a follow- up experiment. Because of the many PCR cycles required in the ADSE scheme, the amplification inevitably leads to the formation of artifacts in the form of longer sequences Tolle et al. (2014), a phenomenon that intrinsically sets a limit to the number of generations that can be explored (see Appendix 1).

As detailed below, the ADSE protocol enables observing a non-trivial evolution of the DNAi ecosystem across generations, the emergence of dominating species and non-monotonic population evolution. It also enables to appreciate the role of inter-specific interactions and their contribution to fitness.

Evolution of the DNAi ecosystem

The main output of the synthetic evolution that we are proposing is the dataset {DNAi}j obtained by sequencing DNAi at the various cycles (1 ≤ j ≤ 24 being the index of the cycle). We find that the initial random ecosystem markedly changes with ADSE generations as shown in Figure 2a, in which we have plotted the evolution of: (i) the fraction FD of the total population formed by distinct nucleotide sequences (red dots). FD drops from 1 to nearly 0, indicating that initially DNAi are all different from each other, while after 24 generations the number of distinct sequences is much smaller than the number of DNAi. This indicates that most of the initial sequences become progressively “extinct”; (ii) The fraction F10 of the total population formed by the 10 most abundant DNA sequences (black dots). F10 is initially close to 0 - being each sequence represented by a single DNAi, and grows to about 25% at cycle 12 and 55% at cycle 24 - indicating that the offspring of 10 initial “mother” sequences becomes the majority of the system. This loss of diversity can also be quantified by the zipped file size of the list of sequences Benedetto et al. (2002) or by computing its Shannon entropy Hill (1973), both markedly decreasing during evolution (see Appendix 1).

Evolution of the DNAi population (Oligo1 data). Time is expressed in ADSE cycles. a: fraction FD of the total population formed by different sequences obtained from the experiment (red dots) and computed with the IBEE model (green line), fraction F10 of the total population formed by the 10 most abundant sequences (experimental: black dots, IBEE model: gray line). b: ⟨ω⟩ computed on the whole population in each generation (red dots, left-hand side y-axis); ⟨ ΔGDR⟩ computed on a sample of 1000 randomly chosen DNAi from the population in each generation (blue dots, right-hand side y-axis); data fitting with the IBEE model (green line, left-hand side y-axis). The left and right y axes were scaled so that ⟨ΔGDR⟩ and ⟨ω⟩ computed on a pool of random sequence DNAi would coincide (dashed blue and red lines, respectively). c: boxplots and scatterplots of ⟨ΔGDR⟩ in ensembles of 1000 random sequences (left), 1000 randomly chosen DNAi extracted from the experimental population at cycle 24 (middle) and from the top-10 most populous DNAi at the same cycle (right). The color code is assigned to each point based on its ω value (color bar). d-f: probability distributions p (ω) for cycles 0 (d), 12 (e) and 24 (f). In the latter histogram, empty bins result from sub-sampling. Orange points and lines are the distributions evaluated with the null model. g-i: evolution of the abundance (expressed as fraction of the total population F) of sequences whose ω is small (3 ≤ ω ≤ 5- panel g), medium (6 ≤ ω ≤ 10 - panel h) and large (11 ≤ ω - panel i) as obtained from the experiments (dots) and with the IBEE model (green lines). The model results are averages over 20 simulations. The online version of this article includes the following figure supplements:

Figure 2-figure supplement 1. Time evolution of the fraction of PCR by-products and of their ⟨ω⟩.

Figure 2-figure supplement 2. Probability distributions p (⟨ω⟩)of PCR by-products at cycles 0 and 24.

Figure 2-figure supplement 3. Sequencing technical replicate.

Figure 2-figure supplement 4. Checking PCR effects on DNAi selection.

Figure 2-figure supplement 5. Time evolution of ⟨ω⟩ for the Oligo2.

Figure 2-figure supplement 6. Probability distributions p (⟨ω⟩)at cycles 0 and 18 for the Oligo2.

Figure 2-figure supplement 7. Experimental vs IBEE time evolution of the population zip ratio and of the Shannon Entropy associated to the RSA distribution.

Figure 2-figure supplement 8. Experimental vs IBEE ⟨ω⟩ time evolution, for two different IBEE hyperparameters choices.

Figure 2-figure supplement 9. Experimental vs IBEE ⟨ω⟩time evolution, for different sizes and compositions of the starting population of IBEE model.

Since in ADSE survival depends on bead capture, we expect DNAi-resource binding strength to be an essential ingredient of fitness. To this aim we compute the mean DNAi-resources binding free energy ΔGDR across generations, plotted in Figure 2b (blue squares, right-hand side y- axis). Specifically, ΔGDR has been computed based on DNAi and resource sequences by using the standard “nearest-neighbor approximation” for the thermodynamics of DNA hybridization SantaLucia Jr (1998); Ghosh et al. (2019); Plata et al. (2021), as implemented in the NUPACK tool in Python Zadeh et al. (2011). Being this computation of some complexity, we could perform it only on batches of 1000 DNAi randomly chosen within {DNAi}j, the error bars expressing the uncertainty introduced by such down-sampling (see Material and Methods section). For reference, a DNA 20mer perfectly complementary to the resource would bind to it with an energy of approximately -24 kcal⁄mol.

To perform a faster - and thus more complete - analysis of the resource-binding energy in the evolving population, we have introduced ω, a simpler quantifier than ΔGDR. ω is the length of the longest consecutive number of bases within each DNAi that is complementary to the resource sequence (considering all possible relative positions of the two), in which we allow for single pairing errors and 1-base bulges Mambretti et al. (2022a). ω ignores factors that are relevant for the free energy, such as the specificity of the sequence and the fraction of CG pairs, and thus is inadequate to evaluate the binding strength of specific DNAi. However, its average value ⟨ω⟩ computed for the entire population in each generation (Figure 2b, red dots, left-hand side y-axis) grows almost identically to⟨ ΔGDR⟩ (blue dots, right-hand side y-axis). The ⟨ω⟩ axis has been scaled so that its value (computed on a pool of random sequence DNAi - dashed red line) matches, in the plot, ⟨ΔGDR⟩ (computed on the same pool - dashed blue line). A more detailed comparison between ω and ΔGDR is given in Figure 2c, showing box plots with ΔGDR on the y axis and ω expressed through color code. Both are computed from a random selection of 1000 distinct DNAi from the initial (left hand side box), the final populations (central box), and the top 10 most frequent sequences in the final generation (right hand side box). The result further strengthens the validity, in our statistical context, of ω as a quantifier of the strength of interaction with the resources Mambretti et al. (2022a).

Figure 2b shows that, during ADSE, ⟨ω⟩ tends to saturate at a value ωsat ≈ 10, indicating that no relevant selective advantage is gained when ω > 10, in agreement with the notion that the residence time of hybridized oligomers becomes larger than typical experimental times when ω > 12 Di Leo et al. (2022).

Figure 2d-f describes the evolution of the ecosystem {DNAi}j by showing p (ω), the fraction of DNAi having overlap ω with the resource evaluated in the initial pool (panel d) and at generation 12 (e) and 24 (f). p (ω) clearly evolves, its small ω components being progressively lost, while individuals with large ω grow in number, as they are more successful in being selected and amplified. It might be worth pointing out that the appearance of non-zero p (ω > 13) at generation 12, while p (ω > 13) = 0 in the initial distribution, is an effect of the under-sampling involved in the sequencing procedure.

The different destiny of DNAi with distinct ω values is shown in Figure 2g-i, where we show the temporal (i.e. across generation) evolution of the fraction of the total population whose ω is in the following ranges: 3 ≤ ω ≤ 5 (panel g), 6 ≤ ω ≤ 10 (panel h) and 11 ≤ ω (panel i). As expected, DNAi with weak affinity to the resources decrease, while those with large affinity increase. Interestingly, DNAi with intermediate affinity exhibit a non-monotonic behavior, indicating that the conditions for survival change during the evolution, reflecting the evolution of the ecosystem.

Null Model and Eco-evolutionary Algorithm

In order to better understand the experimental outcomes, we build first a null random model without evolution and then an Individual-Based Eco-Evolutionary (IBEE) model enabling predictions for p (ω).

The null model describes the interaction between individuals in the initial pool {DNAij=0} of random sequences and the resources within a purely combinatorial framework (see Methods and Appendix 3 for mathematical details). In this model, we attach a random string of length 50 to the resource string in a random position, and we compute the maximum consecutive overlap, accepting the binding only if it is at least formed by 3 complementary basis.

The resulting analytical p0(ω), plotted in Figure 2d (orange crosses) closely matches the data obtained from sequencing, confirming the random-sequence nature of {DNAi}j=0}. The analytical p0(ω) extends to a ω range where no data are available due to the limited size of the sequenced pool. The comparison between p0(ω) and p (ω) in panels e and f shows that the exponential decay of p0(ω) at large ω is maintained even in generation 12 and partly in generation 24, further supporting the notion that the selective advantage of ω saturates at large ω.

In the IBEE algorithm we consider Np individuals. Each has a fitness fi(ω) (i = 1, 2, .., Np) that depends on its affinity ω with the resource. At time t = 0 we assign ω to each individual based on p0(ω) from the null model. Then, for each evolutive cycle, we model competition and selection so that out of Np individuals, only Nr survives. This is attained as a combination of two processes: a fraction x of the Nr sequences results from sampling individuals from the Np population, each with a survival probability f (ω) ∈ (0, 1) (selection); the remaining fraction 1 − x is extracted completely from the Np individuals. The latter group is meant to mimic “neutral drift” (as it would be expressed in evolutionary language) provided by non-specific binding to the beads. The Nr survivers are amplified by identical copying back to Np, the starting population of the next evolutionary cycle (the introduction of very rare mutations do not affect the results). We perform 24 evolutionary cycles.

As expected, the outcome of the eco-evolutionary dynamics strongly depends on the shape of the fitness function. We model it as: where ωmax is the (cycle-dependent) largest value of ω within the actual population and, ω* = ω if ω < ωsat, while ω* = ωsat otherwise. ωsat and γ are parameters to be tuned in the comparison with the observed ⟨ω⟩. ωsat expresses the loss of fitness gain for ω > ωsat yielding the saturation of ⟨ω⟩. γ represents the strength with which f ⟨ω⟩depends on ω, and thus the rate at which low ω individuals are discarded. γ may of course depend on time on account of the evolving ecosystem.

After grid search we find x = 0.9, indicating that the random drift contributes to about 10% of survival at each cycle, and ωth = 10, in agreement with the observations prompted by the saturating behavior of ⟨ω⟩. We also find that the data cannot be approximated with a single value of γ, as visible in Figure 2-figure supplement 8. Data can instead be very well matched assuming γ = 3 for the first 5 cycles and γ = 1 for the remaining cycles, as shown in Figure 2b, green line. With the same choice of parameters, the IBEE fitness based model captures the decrease in the diversity of the DNAi ecosystem (Figure 2a, green and grey lines) and the temporal (i.e. across generations) evolution of the relative population abundance of DNAi in the three ω intervals in Figure 2, panels g, h and i.

The effect of the system size and initial conditions on the IBEE results, discussed in the Appendix 2, do not qualitatively change the models results. Error-bars on simulations have been obtained by averaging 20 independent runs, starting from the same initial conditions. As can be observed, the variability among simulations is negligible.

Also, the IBEE model with a fixed γ would respect the increasing, decreasing, and non-monotonic trends of the data in Figure 2, but not quantitatively. The change in γ, and thus in fitness, is indeed necessary to reproduce the observed⟨ω⟩with the IBEE model. This is a key result of our investigation because it indicates that, even in simple conditions of ADSE environment (fixed resource and low mutation rate), survival is controlled by more than affinity to the resources, as discussed in the analysis below.

Self and mutual DNAi interactions are evolutionary drivers

While ω is certainly a key driver of the observed evolution, it is clearly not the only one. The fact that the top 10 most represented sequences become, in the last cycle, about 55% of the total population (Figure 2a) implies that, even among sequences with large ω, only a tiny minority come to dominate, while the largest part of them eventually disappear. Moreover, the 10 most represented sequences do not stand out for their particularly large ΔGDR or ω, as noticeable in Figure 2c. The same data are plotted in Figure 3a as a p (ΔGDR) distribution to enable comparing the free energy distribution in the initial population (purple shading), in the final population (blue columns), and in the top 10 (black columns). These elements support the notion that survival and dominance must also be due to factors in addition to ω, which we thus explored.

Distribution and evolution of free energy quantifiers (Oligo1 data). a: probability distribution for the DNAi-resource binding free energy computed by NUPACK, p⟨ ΔGDR⟩, for the initial population (grey shading), for the final population (random choice of 1000 DNAi - cyan columns, top 10 most populous sequences - black columns). b: time evolution, expressed in cycles, for various mean free energies ⟨ ΔG⟩, normalized to their value computed on pools of random sequences. All ΔG are computed by NUPACK on sets of 1000 individuals: ⟨ΔGDR⟩ (blue dots); unimolecular self-interaction ⟨ΔGself⟩ (green dots); bimolecular mutual DNAi interaction ⟨ΔGDD⟩ (red dots); mutual, self-subtracted interaction ⟨ΔΔG⟩ (yellow dots). c: scatter plot of ΔGself vs. ΔGDR computed for 1000 DNAi in the final population (blue squares). Red dots mark the point relative to the 10 most populous sequences, as identified by the labels. Note that the x-axis scale of panels a and c is the same, enabling identifying sequences. d: scatter plot computed on 104 DNAi pairs from the final population comparing ΔGself,j + ΔGself,l and ΔGDD,kl (green squares). Red dots mark the pair formed by the 10 most populous sequences, some of which identified by labels. With respect to the condition ΔGDD,kl = ΔGself,j + ΔGself,l (black line), data are on average displaced by ΔΔG ∼ 7.5 kcal/mol (yellow arrow). The online version of this article includes the following figure supplements:

Figure 3-figure supplement 1. Three examples of different relative positions for sequence-sequence attachment.

Figure 3-figure supplement 2. Scheme of target-DNAi interaction, from a combinatorial standpoint.

Figure 3-figure supplement 3. Null model without threshold, analytical and simulated, with an even and uneven nucleotides distribution.

Figure 3-figure supplement 4. Null model with/without threshold, simulated and analytical.

Figure 3b shows the value, across the 24 generations, of two other contributions to the total free energy, normalized to their value computed in random sequences, so to enable comparing their relative variations: ⟨ΔGself⟩ (green symbols), the average unimolecular free energy, expressing the average strength of the internal folding of each DNAi; ⟨ΔGDD⟩ (red dots), the total bimolecular DNAi-DNAi free energy, comprising both self and mutual interactions. For comparison, we also plot the normalized value of ⟨ΔGDR⟩ (blue dots). As for the case of ⟨ΔGDR⟩, ⟨ΔGself⟩ and ⟨ΔGDD⟩ are computed by randomly selecting 1000 individuals or pairs, respectively, from the population at cycle j and using the NUPACK tool to compute the values.

Figure 3b reveals that ⟨ΔGself⟩ grows in time even more than ⟨ΔGDR⟩, but with a different progression: ΔGDR grows faster in the first cycles to later saturate, while the growth of ⟨ΔGself⟩ is more uniform.

Since ⟨ΔGDR⟩ is computed as the free energy of the whole DNAi-resource structure, it includes contributions of self energy associated to hairpins in the DNAi. Thus the growth of ⟨ΔGDR⟩ could actually depend on the growth in ⟨ΔGself⟩ (but not the contrary). To investigate this possibility, Figure 3c displays the scatter plot between ⟨ΔGDR⟩ and⟨ ΔGself⟩, for a randomly chosen subset of DNAi at j = 24. In evidence are the 10 points corresponding to the 10 most populous sequences. The plot shows weak or no correlation, and a relevant shift with respect to ⟨ΔGself⟩ = ⟨ΔGDR⟩ (black line), demonstrating negligible dependence of ⟨ΔGDR⟩ on ⟨ΔGself⟩ and indicating that these two quantities reflect two independent driving forces in the ADSE selection mechanism. The existence of two different growth regimes suggests that in the first stages of ADSE the selection is mainly dominated by affinity with the resources, while in later generations the requirement of stronger unimolecular folding becomes more important.

A similar scatter plot analysis for ⟨ΔGDD⟩ yields a different outcome. Figure 3 d compares the ⟨ΔGDD⟩ computed for two selected DNAi (individual “k” and “l”) and the sum of ΔGself of the same two DNAi. The apparent correlation indicates that, on average, a large part of ΔGDD simply embodies the growth of self-energy, although the difference ΔΔG≡ΔGDD,kl − ΔGself,k − ΔGself,l (orange arrow) is non-negligible. Figure 3b shows the behavior of ⟨ΔΔG⟩ across cycles (orange squares). Despite the resulting mild growth of ΔΔG might appear not relevant, it actually indicated that ADSE, independently of ω, selects strings that have higher reciprocal affinity than a random DNAi set (see Appendix 2). Indeed, the selection process could instead have promoted a decrease of the same quantity. It should be noted that mutual interactions might also involve the unfolding of hairpins self-structures, in which case their strength is much larger than ΔΔG.

Self-interactions and mutual interactions can compete with the binding of DNAi to the resources either when they involve the same nucleobases or through steric hindrance. Therefore, we could expect that the increase of ΔGDR would lead to a decrease of both ΔGself and ΔGDD. Their unfore- seen growth in the ADSE process is hence an indication of the selective advantage they convey. We interpret this behavior as an indirect sign that mutual DNAi interactions, more than just an impediment, are major deadly threats for their survival. To avoid them, DNAi need screening. Indeed, the strong growth of self-interactions, and the mild increment in mutual interactions, could represent the emergence of “defensive” strategies, as discussed below.

In our search for evolution quantifiers other than resource binding strength, we also found that selection favors binding to resource sequences close to their resource 3’ terminal, away from the bead surface (see Figure 4-figure supplement 2). While this is expected, being that terminal less constrained and in a less crowded environment, it also provides useful clues to further analysis.

Natural history of DNAi species (Oligo1 data). a: fraction F of {DNAi} that belong to a choice of specific species as a function of the ADSE cycles, in linear (left) and logarithmic (right) scale. Arrows connect the initial condition (one individual per species at cycle 0) to the earliest detection via sequencing, across the 6 order of magnitudes gap (grey shading). The same growth is assumed for species 12.1, suggesting its appearance by mutation occurred at generation 9. b-f: self-interactions (b1, c1, e1), resource interactions (b2, c2, e2, d1, d2) and mutual interactions (b3, c3, e3, d1, d2, f) of selected species, sketched as per the NUPACK output. Nucleobases color coded (G - black, C - blue, A - green, T - red). Paired bases are connected. Double and single stand regions are represented as straight and curved lines, respectively. As in Figure 1a, terminal blocks of DNAi are marked as graphic double helices colored according to the legend of panel a, and beads as sketched yellow spheres. Yellow shading: section of DNAi complementary to resources. Pink frames: regions of hybridization between DNAi. b: interactions involving species 13.0 including its homodimerization (b3). c: interactions involving species 8.0, including its binding to 13.0 (c3). e: interactions involving species 12.1, including its homodimerization (e3). d and f: DNAi heterodimers interactions suggesting parasitism (d1), and possibly mutualism (d2) and mutual damage (f).

The online version of this article includes the following figure supplements:

Figure 4-figure supplement 1. Evolution of the intra-species interaction strengths (⟨ΔGpp⟩).

Figure 4-figure supplement 2. 2D probability for selected DNAi, shown at different experimental cycles, to have a given ⟨ω⟩ jointly with a given attachment coordinate to a target strand.

The evolution of DNAi species

We examined the evolution of DNAi species, i.e. the change in numerosity of the groups of DNAi having equal sequence, which we report in Figure 4a relative to a limited set of (mainly) successful species (for a larger selection see SI). Data in Figure 4a are expressed as the fraction F of the total population that belong to a given species. It can be noticed that species that are later becoming dominant “appear” in our analysis only at generation 5 in a single or a few copies. This is reasonable since we sequence a sample of ∼106 DNAi over the much larger population of ∼1012 survived individuals.

Species are named after their ω value, combined with an index (e.g. “12.0”, “12.1”, “12.2” …) expressing their ranking in populousness. In Figure 3, panels c and d, we marked and labeled in red the free energy values associated to the 10 species dominating in the last generation and labeled the corresponding species or pairs of species. It can be noticed that the dominant species are not - as one could naively expect - at the extremes of the energies distributions and assigned, confirming that the statistical trends do not enable, by themselves, a full understanding of ADSE.

The populousness of ADSE species does not have standard patterns of evolution. This is true even when limiting the observation to the successful species in Figure 4a whose population growths are varied and do not reflect in any simple way the energy quantifiers. Conversely, species with similar energy quantifiers may have opposite fate, either becoming dominant or rapidly extinguishing or else displaying a non-monotonic population evolution. It is worth pointing out that all species start equal, each represented in the original pool by a single molecule, which on this scale would appear as a F ≈ 10−12 (black dot in Figure 4a, right panel).

We thus decided to seek further understanding on the survival of the fittest in ADSE by inspecting the “natural history” of a few species (see Tables 2, 3 in the Supplementary Information).

13.0 and 11.0 (orange and gray dots in Figure 4a) are, respectively, the most and second most populous species in the last generation. It might be relevant to mention that the replica of ADSE described in the SI leads to different dominant sequences, as expected on the basis of the distinct initial pool and on the key role of randomness and sampling in the first cycles. At generation 24, 13.0 and 11.0 have grown to more than 85% and 53% of DNAi with ω = 13 and ω = 11, respectively. 13.0, and 11.0 likewise, have a weak self-interaction (Figure 3c), corresponding to hairpins that provide a mild defense to mutual interactions, as schematically sketched in Figure 4, panel b1. The hairpins involve most of the bases that are complementary to the resource (yellow shading), leaving however a few unpaired bases that might act as toehold to initiate resource binding. The resource binding takes place at its 3’ terminus (panel b2). A bit puzzled about the success of these two species we also explored their capacity of mutual interactions and found that both of them are capable of forming homodimers, as shown in panel b3 for 13.0 (binding energy marked in Figure 3d). Such dimers can bind resources at both ends (yellow shading regions), giving 13.0 and 11.0 self-screening and divalent binding capacity. We argue this to be a crucial drive toward success of these two species, which also explains why their growth has such an increment in the latest generations, following the probability of dimer formation.

The third most populous sequence (species 8.0) is characterized by a weak resource interaction which includes a bulge (see Figure 3c). While its survival rationale appears clear - good defensive self-binding (panel c1) and interaction targeting the 3’ end of the resource (panel c2) - it is hard to accept this could be the cause of the success. Again, inspecting the mutual interaction we found that 8.0 efficiently interacts with 13.0 (as marked in Figure 3d), in a region (frame in pink) that does not conflict with the capacity of 13.0 to bind the resource (panel c3). We speculate that this parasitic capacity of 8.0 adds to the weak intrinsic binding to lead to a remarkable success of this species.

As a test to this concept, we focus on species 5.1, whose binding strength to the resource is the weakest in the top 10 species and one of the weakest among all the survivors of generation 24 (Figure 3c), a feature that questions its success. However, by inspecting mutual interactions, we find its binding to 11.0 to be strong and stable (pink frame in panel d1 and Figure 4d1). An analogous behavior is found for species 5.0 and its interaction to 12.0 (panel d2), confirming that parasitism is an emerging successful survival solution in ADSE.

We then inspect 12.1 since it has largest ΔGDR and ΔGself among the top 10 species (Figure 3c). Species 12.1 has all the features to succeed in ADSE: it forms a weak hairpin with the bases involved in the binding with the resources and a strong hairpin that protects the rest of the ssDNA segment (panel e1); it strongly binds to the resource with the bond ending at its 3’ terminal (panel e2); it forms homodimers capable of double resource binding, analogous to that of 13.0 (panel e3). Remarkably, 12.1 emerges only late in ADSE (Figure 4a, blue dots), a very unusual behavior when compared to the other many species we have considered. The combination of this late appearance and of the remarkable growth in F afterward suggests that 12.1 is the outcome of one of the rare mutations that can occur even with the high quality PCR that we adopted. By assuming an initial growth analogous to that of dominant species (black arrows in panel a), we argue that such mutation occurred around generation 9, where we have found ancestors having sequence equal to 12.1 except for 4 bases.

F has a monotonic increase for most of the species that become dominant in the latest generations. This behavior is however not at all general. Non-monotonic F, growing during in the first generations and decreasing to extinction afterward, is the general behavior for the majority of species that do not become extinct in the first cycle, as also shown - through their average behavior - in Figure 2h. An examples of these is species 7.X (empty dots in Figures 4a and 4b).

More intriguing is when a non-monotonic behavior is observed for one of the dominating species. This is the case of 11.1, whose growth can be again attributed to an effective screening and a good binding to the resources, but whose decrease is harder to justify. To understand it we again investigate mutual interactions, and we find a strong 11.0-11.1 bond (sketched in panel f), which, differently from the DNAi-DNAi complex examined above, competes with resource binding. We thus speculate that the fall of 11.1 is a consequence of the rise of 11.0, which has become frequent enough to make the formation of the complex 11.0-11.1 probable, which lowers the survival probability of both species. This could also explain why species 11.0 has such an irregular growth pattern, with a slowing down when 11.1 becomes highly populated.

The insight gained by these case-studies enables appreciating how self and mutual interactions can affect the fate of species in ADSE beyond what we could discern through distributions and average quantifiers. Indeed, this analysis shows that the success of the dominant ASDE species can be achieved through different combinations of resource and mutual DNAi interactions, a fact that makes the evolution of population so varied and generation dependent and explains why, in the latest generations of ADSE, the population distribution p(ΔG) (Figure 3a) is so irregular.

Discussion and Conclusion

We have here introduced ADSE, a synthetic molecular evolution protocol that exploits sequencing technology and DNA interaction computability to provide a test-bed for key concepts in ecology and evolution. Although the simple scheme of ADSE enables to perform studies in which mutations and resource drifts can be introduced, we performed experiments by adopting the most simple environment within this protocol, i.e. by holding fixed the capture sequence and minimizing mutations. This choice was aimed at achieving a condition dominated by competition and selection, which enables investigating the nature of fitness in this simplified evolution process.

Our experiments and their comparison with theoretical models indicate that fitness is not a simple function of the direct competitive advantage of strong binding to the resources, as it could have been naively expected. Resource binding is indeed the dominant factor in the first part of the evolution (cycles 1-5), producing a fast growth of ⟨ω⟩ compatible with a strong dependence of the survival probability on ω (f ∝ω3). However, as ⟨ω⟩reaches a value corresponding to bonds of moderate stability Woodside et al. (2006) - approximately from generation 6 onward - the selective pressure related to resource binding decreases (f ∝ ω), a condition enabling to appreciate other factors at play. Which are these factors is suggested by the different growth patterns of ΔGDR and ΔGself, the former dominating the first cycles while the latter drives the later stages. However, the quantification of binding energies is not sufficient to predict the fate of individual species, that also depends on interaction details (location of binding and hairpins, secondary structures) and possibly on kinetics.

In fact, by design, DNAi can interact with themselves, both internally - forming hairpin-like struc- tures - and mutually - leading to the formation of complexes. These interactions can either coexist when they involve non-overlapping sequences - or they can become competitive. Our experiments reveal that, among the wide variety of conditions made available by the initial random seed, ADSE generally promotes those that combine a good interaction with the resources and a structure capable of screening from mutual DNAi interactions conflicting with binding to the beads. By exploring hybridisation-dependent DNAi secondary structures - the only ones accessible with simple analytical tools, we found three basic motifs: hairpins due to self-interactions (as for species 12.1), formation of DNAi homoduplets (as for species 13.0 and 11.0), and formation of dimers of DNAi belonging to distinct species. The latter can bring to distinct prototypical behavior: it can be the basis of parasitism, as in the case of the pairs 8.0-13.0 and 5.1-11.0, it can lead to extinction with no benefit for either species (as in the case of 15.0-12.0), or it can provide - at least in principle mutualism. While we do not have a solid proof, clues suggest that cooperation might be active in the case of the interaction between species 5.0 and 12.0, since 12.0 has a weak self-defensive system and since the populations of 12.0 and 5.0 evolve very similarly (see Table 2). Finally, we cannot exclude a possible role of the formation of multi-DNAi multivalent complexes. This is the case of species 12.0 and 15.0 whose pattern of multivalent mutual interactions can, in principle, allow higher order interactions (see Appendix 2).

Fitness in ADSE is the outcome of a complex interplay of all these elements. The history of single species in Figure 4 indicates that the nature of fitness is beyond what can be captured by analyzing the probability distributions of the relevant parameters describing structure and interactions. Moreover, the variety of population evolution patterns enlightens the fundamental fact that despite the constancy of resources, the progressive modification of the ecosystem brings about a change in the competition modes, and thus in fitness. A forthcoming work focused on a limited number of species will be devoted to better disentangle the nature of fitness in ADSE.

The evolution of the ADSE ecosystem, which corresponds to a marked decrement of its entropy with 2/3 of the initial species become extinct in the first 5 generations - and potentially terminating with the indication of a single winner species leads instead, in the last cycles, to a significant number of dominant species all still growing at the expense of subservient ones, indicating that in their direct competition none of them is strongly prevailing despite - or maybe thanks to - their distinct survival strategies and their very different population share. Hence, while the niche hypothesis could still be verified in the long run - beyond the experimental limitations due to PCR - it is clear that its drive is weak, suggesting that it could be overcome by environmental fluctuations, thus allowing for coexistence of different species even in a single niche environment.

Methods

Our experimental design takes advantage of a selective capture mechanism where magnetic beads carrying single-stranded DNA filaments of fixed length and sequence (resources) target DNA individuals (DNAi) (Figure 1-figure supplement 1) present in a DNA library based on their level of complementarity. This process of selection is carried out through subsequent steps that are described in detail in the next paragraphs and represented in Figure 1-figure supplement 2.

Library design

The DNA library contains 100-nt-long sequences where a randomized central region of 50 nucleotides is flanked by 25-nt-long fixed sequences at both its 5’ and 3’ ends. The fixed regions provide an anchor point for primer annealing, required to perform PCR reactions during the amplification phase (see below). These terminal segments are made inactive by hybridization with oligomers of perfect complementarity (blockers), so that they are not involved in the selection phase. The blockers, as well as the fixed regions of the DNA library, have been designed to avoid hybridization with the resources. In addition, the blockers carry a phosphate group at their 3′ end to prevent them from functioning as primers during the PCR amplification. Following the above described criteria, we designed two sets of sequences (Table 1 in the Supplementary Information), called Oligo1 (whose results are presented in the main text) and Oligo2 that was used as a replication experiment. All the oligonucleotides used in this work were purchased from Integrated DNA Technologies, Coralville, IA, USA (Table 1).

Beads preparation

The capture of DNAi within the DNA library was performed with carboxylic acid magnetic beads (M-270 Dynabeads, Invitrogen, Carlsbad, CA) coated with the resources. The resources are 20- nucleotide-long, 5’-amino-modified oligonucleotides. Their coupling to the beads surface was performed according to the manufacturer’s instructions. Following the activation and coupling procedure, the beads were washed in Tris-HCl (50mM, pH 7.4), and stored in the same buffer in single-use aliquotes.

Sample preparation

The starting samples were prepared in 1X SSC buffer (0.15M sodium chloride, 15mM sodium citrate, pH 7.0). In detail, the Oligo1 (or Oligo2) library (0.75nmol) was mixed with the blockers (2.25nmol) in 1:3 molar ratio to saturate all the available interaction sites between the blockers and the fixed regions of the DNA library. The sample was then denatured at 95°C for 5 minutes, then slowly brought to room temperature using a thermal cycler (MasterCycler Nexus Gradient, Eppendorf, Hamburg, Germany).

Selection phase

The beads, coupled with the resources, are mixed with the sample, prepared as described above. The capture of the DNAi is carried out at 40°C for 2 hours in stirring (600rpm, ThermoMixer, Eppendorf). Once the selection phase is completed, the sample is incubated for 2 minutes on a magnet and the supernatant is removed. The beads, that are now bound to the captured DNAi, are washed three times in SSC 1X buffer to eliminate the aspecific sequences. Finally, the beads are resuspended in water and incubated for 5 minutes at 60°C to recover the DNAi from the resources. The sample is quickly placed on the magnet and after 2 minutes the supernatant containing the selected DNAi is collected and then quantified by NanoDrop (Thermofisher Scientific).

Control experiments were performed showing that no artifact in ADSE are introduced in ADSE because of non-specific interactions with the magnetic beads (see Appendix 1, Experimental controls).

Amplification phase

The captured DNAi are then amplified by PCR with Q5® Hot Start High-Fidelity DNA Polymerase (New England Biolabs, Ipswich, MA, USA). DNA samples were diluted 10 times and 3μl were used for each PCR reaction. The PCR was performed in a final volume of 25μL using 0.25μL of polymerase, and each primer had a final primer concentration of 2.5μM. To allow the regeneration of the single strand library, the reverse primer was designed to carry a 5′ biotin modification (see next paragraph for details). The thermal protocol was the following: i) denaturation step at 98 ° C for 2 minutes, ii) 28 cycles characterized by three thermal steps: 10 seconds at 98°C, 10 seconds at 68°C for Oligo1 (69°C for Oligo2) and 1 second at 72°C, iii) 2 minutes at 72°C. The annealing temperature was kept higher than conventional protocols to endure that hairpins or DNAi dimers are melted. For each generation, 10 different PCR reactions were performed. The PCR products were then checked for size on a 2% TBE 1X agarose gel.

Control experiments were performed showing that no artifact are introduced in ADSE from the amplification steps (see Appendix 1, Experimental controls).

Regeneration phase

PCR products were purified by precipitation with 2.5 volumes of ethanol and 0.1 volumes of sodium acetate 2M, pH 5.2. After an overnight precipitation at -20°C, the pellet was washed with ethanol 75%, resuspended in water and quantified using the Qubit fluorometer (Invitrogen, Waltham, MA, USA). The single-strand regeneration was performed with streptavidin-coated magnetic beads (M- 270 Dynabeads, Invitrogen, Carlsbad, CA) according to manufacturer’s instructions. Briefly, the PCR product was incubated for 15 minutes on a rotator with the magnetic beads. After 2 washes with binding and washing buffer (10mM Tris–HCl pH 7.5, 1mM EDTA, 0.2M NaCl) and a third one with SSC 1X buffer, an alkaline denaturation was performed by incubating the beads with 150mM NaOH for 10 minutes to induce the separation of the two DNA strands. This way it is possible to recover the unlabeled DNA strand in solution. The ssDNA is then collected and buffered with 1.25M acetic acid and TE 1X. The regenerated ssDNA is ready for the next round of selection and amplification.

Sequencing of the recovered products

To check the growth of DNAi species in our experimental model, some generations were sequenced through next-generation sequencing techniques. To prepare the sample for sequencing, the DNAi species captured as described in the “Selection phase” paragraph were first PCR-amplified using the same conditions discussed before. However, in this case, both forward and reverse primers were unlabelled. After PCR purification, performed as already described (see “regeneration phase” paragraph), the products were quantified and used to obtain libraries. About 300ng of DNA were used as starting material for the NEBNext® Ultra™ II DNA Library Prep Kit for Illumina ® (New England Biolabs), and libraries were prepared following the manufacturer’s instruction. Sequencing was performed with the NextSeq™ 550 sequencer (Illumina, San Diego, CA, USA) and a paired-end strategy to obtain 75-nt long reads.

A technical replica was performed to assess the fluctuations intrinsic to sequencing in the context of ADSE (see Appendix 1, Experimental controls)

Replicas

The experimental workflow was repeated on two different sets of DNA libraries: Oligo1, whose results are described in the main text, and Oligo2, used as a replicate. The experimental conditions and procedures were the same for both sets of oligonucleotides, the only differences being the sequences of the fixed parts of the DNA libraries (and hence of the blockers and primers) and of the resources (Table 1). As a consequence, also the PCR conditions had to be adjusted, in terms of the annealing temperature that is one degree higher for Oligo2.

Eco-Evolutionary algorithm

To support experimental observation with a simple abstract model, we developed an Evolutionary Algorithm where a population of Np sequences evolves in presence of Nr < Np shorter sequences. In particular, Nr individuals in this population are selected at each cycle depending on their fitness, i.e. on their affinity to the resource, expressed via their ω with it. These survived sequences are then amplified by a factor of (roughly) Np/Nr. In this work, two types of fitness functions have been explored: the first one is merely linearly proportional to the ω of each individual (i.e., where Σω is the sum of the ω values of the whole population), while the other one is a modification of the previous fitness which sets it to be beyond a threshold value ωth. The code is written in C++, exploiting MPI Message Passing Interface Forum (2021) and Armadillo Sanderson and Curtin (2016, 2018) libraries for acceleration.

NUPACK calculations

We resorted to NUPACK for nucleotide sequences analysis, for the prediction of their free energies at equilibrium, either alone or when binding to one or two other oligomers. Schemes like those shown in Figure 4 have been obtained via the NUPACK web application Zadeh et al. (2011), while massive ΔG calculations have been performed with custom Python codes exploiting NU- PACK Python package (v4.0.0.27) Fornace et al. (2020); Dirks et al. (2007). The model is dna04 and ensemble=‘some-nupack3’; T = 40°C and [Na+] = 0.24 M, as in the experiments. Concentration of each species in the tube has been arbitrarily set to 10−6 M. Detection of hairpins and other secondary structures has been performed visually (thanks to the oxView software Bohlin et al. (2022); Poppleton et al. (2020)).

Supporting information